---
title: "leaflet_in_R"
output: html_document
date: "2024-01-18"
---

```{r setup, include=FALSE}


list.of.packages <- c("leaflet", "htmltools", "rgdal")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) install.packages(new.packages)

library("leaflet")
library("htmltools")
library("rgdal")

```

# Leaflet

Leaflet is one of the most popular open-source JavaScript libraries for interactive maps. It's used by websites ranging from The New York Times and The Washington Post to GitHub and Flickr, as well as GIS specialists like OpenStreetMap, Mapbox, and CartoDB (<https://rstudio.github.io/leaflet/>).

```{r cars}

leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=1.73951, lat=52.45895, popup="Where I go to work")



```

## Basemaps

You can also embed plots, for example:

```{r pressure, echo=FALSE}
leaflet() %>% addProviderTiles(providers$OpenTopoMap) %>% 
  setView(lng = 24, lat = 40, zoom = 03)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Marks and circle markers

We will use the built in quakes dataframe in this example, if you want to explore the dataframe, run this code: summary(quakes). 

It has lat and lon, magnitutde and depth of earth quakes.

### ~ line and calling data!

```{r pressure, echo=FALSE}
leaflet() %>%
  addTiles() %>% 
  addMarkers(data = quakes,
             lng = ~long,
             lat = ~lat)

leaflet() %>%
  addTiles() %>% 
  addCircleMarkers(data = quakes,
                   lng = ~long,
                   lat = ~lat,
                   color = "blue")
```

Clustering labels

```{r, echo=FALSE}
leaflet() %>%
  addTiles() %>% 
  addMarkers(data = quakes,
             lng = ~long,
             lat = ~lat,
             clusterOptions = markerClusterOptions())


```

### Adding labels 

We can also add labels so our user can see information as well as location. 

This is done by using the label function.

We can also adjust many aesthetic parameters or we adjust the aesthetics to show us certain information. So below maybe we only want to see quakes with a magnitutde larger than 5. 

You can also see we have removed the outside ring (the stroke) and changed the size. 
For example: 


```{r, echo=FALSE}

leaflet() %>%
  addTiles() %>% 
  addCircleMarkers(data = quakes,
                   lng = ~long,
                   lat = ~lat,
                   color = ~ifelse(mag < 5, "green", "red"),
                   stroke = FALSE,
                   radius = 4,
                   # label = ~mag,
                   label = ~paste("Magnitude = ", mag, "| Depth = ", depth),
                   )

```


Here we can adjust the size of the marker to show us how big the earthquakes were! (not best example)

```{r, echo=FALSE}
leaflet() %>%
  addTiles() %>% 
  addCircleMarkers(data = quakes,
                   lng = ~long,
                   lat = ~lat,
                   color = ~ifelse(mag < 5, "green", "red"),
                   stroke = FALSE,
                   radius = ~mag,
                   label = ~paste("Magnitude = ", mag, "| Depth = ", depth),
                   )
```

## Pop ups

These allow a bit more information than a label

```{r, echo=FALSE}

quakes_sub = quakes[seq(1, nrow(quakes), 40), ]

leaflet() %>% 
  addTiles %>%
  addCircleMarkers(data = quakes_sub,
                   lng = ~long,
                   lat = ~lat,
                   color = ~ifelse(mag < 5, "green", "red"),
                   stroke = FALSE,
                   popup = ~paste("Wowee this quake had a magnitutde of", htmlEscape(mag), "with a depth of", depth, "meters.", "Holy moley!"))
  


```

## Multiple data and groups

If we have different the layers the user can turn them off and on.

```{r}

small_quakes <- quakes %>% 
  dplyr::filter(mag < 5)

big_quakes <- quakes %>% 
  dplyr::filter(mag > 5)

leaflet() %>% 
  addTiles %>%
    addCircleMarkers(data = small_quakes,
                   lng = ~long,
                   lat = ~lat, 
                   color = "green",
                   stroke = F, 
                   fillOpacity = 0.5, 
                   radius = 5, 
                   label = ~paste("Magnitude = ", mag, "| Depth = ", depth),
                   group = "small quakes",
                   ) %>% 
  addCircleMarkers(data = big_quakes,
                   lng = ~long,
                   lat = ~lat, 
                   color = "red",
                   stroke = F, 
                   fillOpacity = 0.5, 
                   radius = 8, 
                   label = ~paste("Magnitude = ", mag, "| Depth = ", depth),
                   group = "big quakes",
                   ) %>% 
  addLayersControl(
    overlayGroups = c("small quakes", "big quakes"),
    options = layersControlOptions(collapsed = FALSE)
  )
  

  


```

## Shape files and spatial data

```{r}
nycounties <- rgdal::readOGR("https://rstudio.github.io/leaflet/json/nycounties.geojson") 

pal <- colorNumeric("viridis", NULL)

leaflet(nycounties) %>%
  addTiles() %>%
  addPolygons(stroke = FALSE,
              smoothFactor = 0.3,
              fillOpacity = 1,
              fillColor = ~pal(log10(pop)),
              label = ~paste0(county, ": ", formatC(pop, big.mark = ","))) %>%
  addLegend(pal = pal,
            values = ~log10(pop),
            opacity = 1.0,
            labFormat = labelFormat(transform = function(x) round(10^x)))

```

https://r-charts.com/spatial/interactive-maps-leaflet/






