---
title: "Week 2 - Lecture 1 - ggplot"
author: "Sam Speak"
date: "2024-01-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lecture 1 - ggplot

##### Load in packages

```{r install packages}
#install.packages("ggplot2")
#install.packages("ggThemeAssist")
#install.packages("ggpubr")
#install.packages("wesanderson")
#install.packages("ggpattern")
#install.packages("ggVennDiagram")
#install.packages("ggcorrplot")
#install.packages("ggforce")
#install.packages("cowplot")
#install.packages("magick")

```

### Quick recap

We covered a lot in the last week of the course, but one of the main things that we did was introduce you to two different ways of plotting data, using base R and using ggplot.

We talked about how we can create plots with ggplot, how we have to express the aesthetics of a plot with `aes()` and how we can add labels with `lab`.

We also showed you the useful tool `ggThemeAssist()` which is an interactive way to edit figures.

```{r}
library(ggplot2)
library(ggThemeAssist)

ggplot(data = iris) + # define data, we could even add x and y definitions here
  # changing point colour AND symbol based on factor
  geom_point(aes(x = Petal.Length, y = Petal.Width, colour = Species, pch = Species))+
  labs(title = "Petal Width as a function of petal length",
       x = "Petal length (cm)",
       y = "Petal width (cm)") +
  theme_bw()+ # NOTE: here the order matters! We set the overall theme first then overwrite parts of it
  theme(axis.title = element_text(size=10, face="bold", color = "black"))

```

### Adding trends

Quite often we don't want to just display a scatter of points on a plot we need more information.

A good way to determine trends is by using the `geom_smooth()` function to add a trend line.

```{r }
ggplot(data = iris) +
  geom_point(aes(x = Petal.Length, y = Petal.Width, colour = Species, pch = Species))+
  labs(title = "Petal Width as a function of petal length",
       x = "Petal length (cm)",
       y = "Petal width (cm)") +
  geom_smooth(aes(x = Petal.Length, y = Petal.Width, colour = Species),method = glm) +
  theme_bw()+ 
  theme(axis.title = element_text(size=10, face="bold", color = "black"))

```

We can turn off the standard errors and also edit the statistical method used.

```{r }
ggplot(data = iris) +
  geom_point(aes(x = Petal.Length, y = Petal.Width, colour = Species, pch = Species))+
  labs(title = "Petal Width as a function of petal length",
       x = "Petal length (cm)",
       y = "Petal width (cm)") +
  geom_smooth(aes(x = Petal.Length, y = Petal.Width, colour = Species), se = FALSE , method = loess) +
  theme_bw()+ 
  theme(axis.title = element_text(size=10, face="bold", color = "black"))

```

### Plotting distribution

Quite often we don't want to plot just points but distributions.

There are some functions in ggplot for this `geom_boxplot()` and `geom_violin()`

```{r }

ggplot(data = iris) +
  geom_violin(aes(y = Petal.Length, x = Species, fill = Species)) +
  labs(title = "Petal Width as a function of petal length",
       x = "Species",
       y = "Petal length (cm)") +
  theme_bw() + 
  theme(axis.title = element_text(size=10, face="bold", color = "black"))
```

### Adding pattern

There are a wide range of packages available that work with `ggplot()` an example would be `ggpatterns` which provides patterned fills for every ggplot geom that has fill regions.

```{r }
#install.packages("ggpattern")
library(ggpattern)

ggplot(data = iris) +
  geom_violin_pattern(aes(y = Petal.Length, x = Species, fill = Species, pattern = Species)) +
  labs(title = "Petal Width as a function of petal length",
       x = "Species",
       y = "Petal length (cm)") +
  theme_bw() + 
  theme(axis.title = element_text(size=10, face="bold", color = "black"))

```

## Colour palettes

As we have discussed last week you can change the colour of your plots by either typing the name of a colour or by also using a `hex code`.

You can search a specific colours code in Google. A useful website is <https://encycolorpedia.com>.

We can also import some colour palettes that are pre-made for us.

One example of this is all of the Wes Anderson palettes.

```{r }
#install.packages("wesanderson")
library(wesanderson)

wes_palettes
```

```{r }
#install.packages("wesanderson")
library(wesanderson)

life_aquatic <- wes_palettes$Zissou1

ggplot(data = iris) +
  geom_violin(aes(y = Petal.Length, x = Species, fill = Species)) +
  labs(title = "Petal Width as a function of petal length",
       x = "Species",
       y = "Petal length (cm)") +
   scale_fill_manual(values = c(life_aquatic[c(1,3,5)])) +
  theme_bw() + 
  theme(axis.title = element_text(size=10, face="bold", color = "black"))

```

My advice would be that as good as some of these are use them sparingly and you should always consider the application of your figure and the accessibility.

```{r }

cbpalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#CC79A7")

```

## Some thoughts on Theme

```{r }
#install.packages("ggpubr")
library(ggpubr)

Petal_plot <- ggplot(data = iris) +
  geom_point(aes(x = Petal.Length, y = Petal.Width, colour = Species, pch = Species))+
  labs(title = "Petal Width as a function of petal length",
       x = "Petal length (cm)",
       y = "Petal width (cm)") 

Petal_plot +  theme_classic2()+ 
  theme(axis.title = element_text(size=10, face="bold", color = "black"))

Petal_plot +  theme_pubclean()+ 
  theme(axis.title = element_text(size=10, face="bold", color = "black"))

Petal_plot +  theme_light()+ 
  theme(axis.title = element_text(size=10, face="bold", color = "black"))

```

## ggpubr

To make quick and clean looking plots we can use an extension to `ggplot` called `ggpubr`. This provides quick and customised plots that are on the way to be publication ready.

```{r }
#install.packages("ggpubr")
library(ggpubr)

df <-  data.frame(group = c("ARIES","CEFAS","Demonstrators"),
            Lunches = c(13,15,6))

ggdonutchart(df,x = "Lunches", label = "group", fill = "group")

ggdensity(iris, x = "Petal.Length", fill = "Species")

ggviolin(iris, y = "Petal.Length", x = "Species", fill = "Species")
```

### Multiple plots

When we want to create a plot quite often we don't want to show just one figure we want to see multiple to understand the data better

#### Facet_wrap

One quick way we can compare and split apart a plot is by using the `facet_wrap()` function.

```{r }
#### To demonstrate here is a basic histogram of Petal Length

ggplot(iris) +
  geom_histogram(aes(x = Petal.Length)) +
  scale_x_continuous(name = "Petal Length (cm)") +
  scale_y_continuous(name = "Frequency")  +
  theme_pubclean()
```

But what if we wanted to get a histogram of each species respectively to compare them?

```{r }

ggplot(iris) +
  geom_histogram(aes(x = Petal.Length)) +
  scale_x_continuous(name = "Petal Length (cm)") +
  scale_y_continuous(name = "Frequency")  +
  facet_wrap(~ Species) +
  theme_pubclean()
```

#### ggarrange

We can now create many types of plots but what if we want to compare different types of plots in one figure?

We can use the function `ggarrange()` which is part of the `ggpubr` package we looked at before.

```{r }
library(ggpubr)
library(wesanderson)

pallet <- wes_palettes$FantasticFox1

### Make multiple different plots
iris_boxplot <- ggboxplot(iris, x = "Species", y = "Sepal.Length",
    color = "Species", palette = pallet)

iris_points_plot <- ggdotplot(iris, x = "Species", y = "Sepal.Length",
    color = "Species", palette = pallet)

iris_density <- ggdensity(iris, x = "Sepal.Length", 
                          fill = "Species", palette = pallet  )


### Then create a multi panel plot
ggarrange(iris_boxplot, iris_points_plot, iris_density, ncol = 2, nrow = 2)
```

```{r}
### We can combine multiple ggarrange plots together 

iris_top_plot <- ggarrange(iris_boxplot, iris_points_plot, ncol = 2, nrow = 1)

ggarrange(iris_top_plot, iris_density, ncol = 1, nrow = 2)
```

We may want to share the legends of multiple plots to make them more read able

```{r}
ggarrange(iris_boxplot, iris_points_plot, ncol = 2, nrow = 1, common.legend = TRUE)
```

We can also adjust the size of each plot relative to each other

```{r}

iris_scatter <- ggscatter(iris, x = "Sepal.Width",
                       y = "Sepal.Length",
                       group = "Species",
                       color = "Species",
                       palette = pallet)
ggarrange(iris_scatter,iris_boxplot,ncol = 2, nrow = 1, common.legend = TRUE, widths = c(3,1))
```

### Zooming into a plot

```{r }
library(ggforce)

iris_scatter <- ggscatter(iris, x = "Petal.Width",
                       y = "Petal.Length",
                       group = "Species",
                       color = "Species",
                       palette = pallet)
iris_scatter

iris_scatter + facet_zoom(x = Species == "versicolor")
```

### How to add images to plot

We could want to import an image into a plot that we have created. To do this we need to load in the image that we want and the packages `cowplot` and `magic`.

```{r }
library(cowplot)
library(magick)

iris_density <-  ggplot(iris) + geom_density(aes(x= Sepal.Length, fill = Species)) + 
  scale_fill_manual(values = pallet) + theme_cowplot()

logo_file <- file.path(R.home(), "doc", "html", "Rlogo.pdf")

image_behind_plot <- ggdraw() +
  draw_image(logo_file) +
  draw_plot(iris_density)

image_behind_plot

#### But we can alter where we are putting the image 

cow_logo_file <- system.file("extdata", "logo.png", package = "cowplot")

image_on_plot <- ggdraw() +
  draw_image(cow_logo_file,  x = 0.3, y = 0.4, scale = .2) +
  draw_plot(iris_density)

image_on_plot

```

### Some interesting plot types

#### Venn diagrams

```{r }
library(ggVennDiagram)

cbpalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#CC79A7")

fish_data <- paste("fish",1:1000,sep="")
set.seed(20231214)
sample_data <- list(A=sample(fish_data,300),
          B=sample(fish_data,525),
          C=sample(fish_data,440),
          D=sample(fish_data,350),
          E=sample(fish_data,374))

ggVennDiagram(sample_data) + scale_fill_gradient(low=cbpalette[5],high = cbpalette[2])


```

#### Correlation plots

```{r }
library(ggcorrplot)
corr <- round(cor(mtcars), 1)

ggcorrplot(corr)

```

We can alter many aspects of the plot as well as to identify which correlations are not significant.

```{r }
p.mat <- cor_pmat(mtcars)

ggcorrplot(corr, type = "lower", method = "circle",hc.order = TRUE, outline.col = "white" ,p.mat = p.mat)
```

## Exercises

```{r }

```

Useful links:

Searching colours: <https://encycolorpedia.com>

Changing colours: <http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually>

Colour blind accessibility: <https://stackoverflow.com/questions/57153428/r-plot-color-combinations-that-are-colorblind-accessible>

ggplot basics: <https://www.datanovia.com/en/lessons/introduction-to-ggplot2/> <https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html>

ggpubr: <https://rpkgs.datanovia.com/ggpubr/>
