---
title: "Basic Data Wrangling"
author: "Kate Collingridge, Kate Dewally and Kathrin Vossen"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output: 
  html_document:
    theme: flatly
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE) # This tells it to print the code chunks when knitting to html
knitr::opts_chunk$set(eval = FALSE) # This tells it not to run the code chunks when knitting to html
Sys.setenv(tz="UTC") #set environment time zone to utc

```

```{css, echo=FALSE, eval = TRUE}

.lightgrey {

background-color:#F6F6F6;

color: #000; font-size:

16px; text-align:left;

font-family: Helvetica;

vertical-align: middle;

border-radius: 5px;

padding: 20px;

80px

}


body{
  font-family: Helvetica;
  font-size: 14pt;
}

```

::: lightgrey
**In this session we will:**

- Take a look at functions and packages that will help you do some basic data exploration. 

- Investigate some example data

- Learn how to pivot data, how to merge two datasets and how to create subsets of data. 

All the data we will be using are example datasets that are built into R, so there is no need to import external datasets.
:::

<br>
First, load the following packages:

```{r packages}
library(tidyverse)
library(reshape2)
library(ViewPipeSteps)
```

<br>

### Investigating your data

Let's start by taking a look at the `iris` dataset. This is a dataset that is built into R, so we can simply load it:

```{r investigate1}
data("iris")

# It can be useful to make a copy of your original dataset. That way, if something goes wrong, you can go back to a previous version of your data without having to reload it into your R session.

iris_ori <- iris

# Let's get a first overview

str(iris)

# We can see that we have 4 columns containing numerical values and one column containing a factor with 3 levels

summary(iris)
head(iris)
tail(iris)

```

Let's gather some more information about our dataset.

```{r investigate2}

ncol(iris) # We have 5 columns
nrow(iris) # We have 150 rows

# What are minimum and maximum values for our numerical columns?

min(iris$Sepal.Length)
max(iris$Sepal.Width)
mean(iris$Petal.Length)

# If we wanted we could calculate the sum of a column:

sum(iris$Sepal.Length)

## Now we want to take a look at our factors - what are the three levels?

unique(iris$Species)

# We can use this function on any column, but with numerical values we will just get more results:

unique(iris$Sepal.Length)

# Do we have any NAs in our dataset?

any(is.na(iris$Sepal.Length))
table(is.na(iris$Sepal.Length))

# Finally if we want to open the dataset in a separate tab we can run

View(iris)

# You can also do this by clicking on it in the environment section of R studio.
```

<br>

### Making basic changes to your data

We can change things such as column names, or set them to all lower or upper case. This can be useful when we want to merge different datasets together.

```{r change}
# Let's change a column name

names(iris)[names(iris)=="Species"] <- "Type" 
str(iris)

# let's replace dots in column names with underscores

names(iris) <- gsub("\\.", "_", names(iris))
str(iris)

# What if we want to change column names or rows to all upper or all lower case?

names(iris) <- toupper(names(iris))
str(iris)

names(iris)[names(iris)=="TYPE"] <- tolower(names(iris)[names(iris)=="TYPE"])
str(iris)

# We can also change all values in a column
iris$type <- toupper(iris$type)
str(iris)

# Let's reverse these changes for now
iris <- iris_ori

```

Notice the use of `gsub()` in the code above - this is an example of one way that you can manipulate character data. Functions like `gsub`, `grep` and more, allow you to search for a pattern in a string and then either find the position of it, or replace it.

Pattern matching uses regular expressions such as `.` , `*` or `^` which act as wildcards or placement matching tools for the
pattern. Here is an example:

```{r regex, eval=TRUE}
# make a string
example <- c("let us", "find", "a", "pattern")
example

# find a string
grep(pattern = "pattern", example) # found in the 4th element

# replace a string
example2 <- gsub("pattern", "cat", example)
example2

# find a partial string
grep(pattern = ".*ern", example) # found in the 4th element

# replace a partial string
example3 <- gsub(".*ern", "cat", example)
example3
```

We can also use string tools on character columns in a dataset. Say we wanted to add the genus name to the iris dataset

```{r genus}
genus <- "Iris"

iris$Species <- paste(genus, iris$Species, sep = "_")
unique(iris$Species)

# Now we have the full species name

# Suppose we want to separate out the genus and species. We can use the underscore to do this

strsplit(iris$Species, "_")

# This sort of does the job but if we want these in separate columns in our data frame we can use the separate function:

iris %>%
  tidyr::separate(Species, c("Genus", "Species"), "_")

```

There are many pattern matching tools and regular expressions to get your head around - this [cheat
sheet](https://rstudio.github.io/cheatsheets/html/strings.html?_gl=1*1bshcje*_ga*MTg0OTY1NjUzMi4xNjk4MTYwNTAz*_ga_2C0WZ1JHG0*MTcwNDM4NTAxMy41LjAuMTcwNDM4NTAxMy4wLjAuMA..#regular-expressions) has a list of regular expressions and their meanings.

<br>

### Expanding vs reducing your datasets

Now that we have explored the `iris` data a bit, let's try subsetting it based on different conditions. First, let's try making a new dataframe where we only select data for the species "setosa". There are two ways you can do this - using R base functions or using a Tidyverse pipeline.

Let's try a conditional statement using square brackets first:

```{r subset1}
# let's revert back to our original dataframe:
iris <- iris_ori

setosa <- iris[iris$Species == "setosa",]
str(setosa)
unique(setosa$Species) # Note R still knows there are 3 levels for this factor, but the dataset only includes 1
```

You can read this statement as: "Assign an object called 'setosa' using the 'iris' dataset, selecting all rows where the factor Species has the value 'setosa', and selecting all columns". We are selecting **rows** because our conditional statement is *before* the comma.

We can select **columns** if the conditional statement is *after* the comma. For example, we can only select two of the columns:

```{r subset2}
# To select the first three columns:
subset1 <- iris[,1:3]
str(subset1)

# Or to select by column name:
subset2 <- iris[,c("Sepal.Length", "Sepal.Width")]
str(subset2)
```

Now let's try the same using a Tidyverse function. First, this is how to select specific **rows** based on a factor:

```{r subset3}
setosa <- iris %>%
  filter(Species == "setosa")
str(setosa)
```

And this is how to select specific **columns**:

```{r subset4}
# Select first three columns:
subset3 <- iris %>%
  select(1:3)

#Select by column name:
subset4 <- iris %>%
  select(Sepal.Length, Sepal.Width)
```

The next thing we want to try is expanding our dataset by adding a new column, Let's say we want to calculate the petal length to width ratio. We can do this calculation and automatically save this in a new column that we will call "LWRatio":

```{r column1}
iris$LWRatio <- iris$Petal.Length / iris$Petal.Width
```

<br>

### Duplicates

It is important to check your dataset for duplicate data. In the case of the `iris` dataset, we expect to see duplicate entries when we look at individual columns -- we would assume that the same petal length or width will show up multiple times. However, sometimes we need to check if we have true duplicates in our data. We can do this by creating a 'unique identifier' and then using this column to check for duplicates.

```{r duplicates}
# Let's create a unique identifier by pasting together information from the different rows: (KC think we could have a better example here)

iris$UniqueID <- apply(iris[,1:6], 1, paste, collapse = "-")
str(iris)
# now check
table(duplicated(iris$UniqueID))


# We have one duplicate entry, let's check that further

ID <- iris$UniqueID[duplicated(iris$UniqueID) == TRUE]
test <- iris[iris$UniqueID == ID,]
test
# we can remove one of the two entries using the row number:
iris <- iris[-143,]

#An simpler way to do this is by using dplyr package:

iris <- iris_ori

dup <- iris %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup() %>%
  print_pipe_steps()


dup

#Keep only the distinct rows (based on all columns):

iris <- iris %>% distinct()

nrow(iris) # We now have only 149 rows, we have lost the duplicate.


```

### Wide vs long format

As explained in the tidyverse session, there are two main ways to store
datasets - wide and long. Datasets are usually easier to work with in R
in long format, but you may need to switch between them.

Long format is where you have data in an identifying column (or columns)
which repeats, and then your response data all in one column. For
example, let's look at the chick weight dataset.

```{r chickweight}
#  Load the chick weight dataset

ChickWeight <- as.data.frame(ChickWeight)
str(ChickWeight) #You might notice this dataset has some column attributes for time and weight to give them labels and units. You can access these using the attr() function.


# Let's reorder the columns into a more sensible order

ChickWeight <- ChickWeight[, c("Chick", "Diet", "Time", "weight")] # leave the row index blank to keep all rows

View(ChickWeight)
nrow(ChickWeight)
unique(ChickWeight$Chick) # you can also see the Chick column has levels which specify an order for this factor
```

You can see that this dataset has four columns. The first 3 contain
identifying info (which chick is it and which diet is it on) and time of
the measurement. The final column has the chick weight. Thus there are
only 50 different chicks on 4 diets, but as they were weighed at
different times and the data is in long format there are 578 rows in
total. This is a good format for many r functions, but it does repeat
information and is not so easy to read.

The other way to store this data is in wide format, where each chick
just has one row, and each time it is weighed goes in a separate column

```{r wide}
#In tidyverse

ChickWeight %>% tidyr::pivot_wider(names_from = Time, values_from = weight)

#In reshape2

reshape2::dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight")

# You will see we now have each time step in a separate column and one row per chick.
```

There are various terms for converting between these formats. The most
common are `pivot_long` or `melt` for wide\>long, and `pivot_wider` or
`cast` for long\>wide.

Casting a dataset is simple where each unique entity only has one of
each measurement, but sometimes that isn't the case. Maybe it was a bit
difficult to get the chicks to stay still for weighing and you actually
had more than one measurement for each chick in order to take an
average. If you tried to cast such a dataset then you would get an
error, as it can't put more than one weight in each cell in the wide
format table. In this case you might need to specify a function, usually
average.

```{r wide2, error = TRUE}
ChickWeight

#let's say chick 1 was being particularly difficult to weigh on day 6, so we have multiple measurements for that day
ChickWeight <- ChickWeight %>% add_row(Chick = "1", Diet = "1", Time = 6, weight = 66)

ChickWeight_wide <- ChickWeight %>% tidyr::pivot_wider(names_from = Time, values_from = weight)
# See how we get a nonsensical result here, because row 4 and 579 both contain weights for chick 1, diet 1, day 6, and it hasn't been able to do the pivot and has just put the number of observations.

# We need to specify a function to use where there are multiple values

ChickWeight_wide <- ChickWeight %>% tidyr::pivot_wider(names_from = Time, values_from = weight, values_fn = mean)


#We can also do the same thing with dcast

reshape2::dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight")

#In dcast it does run but says "Aggregation function missing: defaulting to length", and you will see all but one cells have a length of 1.
#If we add fun.aggregate = mean the error goes away and where there are duplicates it calculates an average

ChickWeight_wide <- dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight", fun.aggregate = mean)

```

There may also be situations where you want to go from wide to long
format

```{r wide to long}
# In tidyr
ChickWeight_long <- ChickWeight_wide %>% pivot_longer(cols = 3:14, names_to = "Time", values_to = "weight")

# In reshape2
ChickWeight_long <- reshape2::melt(ChickWeight_wide, id = c("Chick", "Diet"), value.name = "weight", variable.name = "Time")
```

Now we have the data back in long form. However you may notice that
there are now 600 observations, when there were originally 578. This is
because each cell has been given a row regardless of whether there was a
measurement in it, so there are 50 chicks \* 12 times = 600 rows. As you
may have noticed in the wide table, several of the chicks weren't
measured every time step (let's imagine they escaped part way through
the experiment), for example chick 16 was only weighed up to day 12. So
by pivoting from wide to long format we have introduced some extra rows.

```{r remove NAs}
ChickWeight_long <- ChickWeight_long[!is.na(ChickWeight_long$weight),]
```

Now we are back to the orginal 578 rows.

### Merging datasets

Before we can merge or join two datasets together, we need to check that
they are compatible. The `janitor` package allows us to compare columns
between two datasets. Let's create an example by setting up two subsets
of the `iris` dataset and renaming one column in one of the subsets to
create a difference:

```{r janitor, eval = TRUE, warning=FALSE, message=FALSE}

# iris <- iris1
subset1 <- iris[iris$Species == "setosa",]
subset2 <- iris[iris$Species == "virginica",]

names(subset2)[names(subset2) == "Species"] <- "Type"

library(janitor)

janitor::compare_df_cols(subset1, subset2)

# We can now see that four of the columns match but species and type do not.
```

To join two datasets together, we have several options:

-   **cbind**: this function merges two dataframes by adding *columns*
    -- so the dataframe is expanded horizontally
-   **rbind**: this function merges two dataframes by adding *rows* --
    so the dataframe is expanded vertically
-   **merge**: base R function within which you can specify different
    ways to merge two dataframes
-   **Tidyverse options**:
    -   inner_join() corresponds to merge(X, Y) -- only retain rows that
        match between X and Y.
    -   left_join() corresponds to merge(X, Y, all.x = T) -- retain all
        rows from X, whether or not they match in Y.
    -   right_join() corresponds to merge(X, Y, all.y = T) -- retain all
        rows from Y, whether or not they match in X.
    -   full_join() corresponds to merge(X, Y, all = T) -- retain all
        rows from both X and Y, whether or not they match.

Source and further info:
<https://bookdown.org/rwnahhas/IntroToR/merge-join.html>

An example:

```{r rbind}
#set the name back to species for subset 2
names(subset2)[names(subset2) == "Type"] <- "Species"

fullDF <- rbind(subset1, subset2)
fullDF

```

Now lets look at some made up data

```{r schools}
# Create some dataframes with random data.

# Random ratings per school and city.
schools <- data.frame(City=c('Lowestoft', 'Lowestoft', 'Beccles', 'Beccles', 'Norwich'),
                      School=c('Pakefield', 'Kirkley', 'John Leman', 'Free School', 'Hewitt'),
                      Rating=c('B', 'D', 'C', 'C', 'B'))

# Town sizes
towns <- data.frame(Town=c('Lowestoft', 'Beccles', 'Norwich'),
                    Population=c(30000, 10000, 270000))

NewRating <- data.frame(NewRating=c('B', 'C', 'B', 'D', 'B'))

# Add the information about population size to the schools table using merge

schools2 <- merge(schools, towns, by.x = "City", by.y = "Town")

# Note we had to specify the column to match by as it was called city in one table and town in another. If we change the column name to be town in both then the merge is simpler as it guesses that you want to join on that column.
names(schools)[names(schools) == "City"] <- "Town"

schools2 <- merge(schools, towns)

# Suppose the schools have been rated again and we want to add the new ratings, we can use cbind

schools3 <- cbind(schools2, NewRating)

```

#### Sorting

Sometimes we need to sort a data frame. Say we wanted to have the town
names in alphabetical order

```{r sort}
data(mtcars)
newdata <- mtcars[order(mtcars$mpg),]

schools3[order(schools3$Town),]
# Or we can order by rating in descending order
schools3[order(schools3$NewRating, decreasing = TRUE),]

```

### Dealing with dates and times

Most data will come with a date and or time, but the formats can vary
widely so it is important to understand how to deal with these in R.
["If you think dates and times are simple you haven't thought about it
enough!"](https://gist.github.com/timvisee/fcda9bbdff88d45cc9061606b4b923ca)

#### Time zones

The first thing to get right is the time zone! You may have noticed that
in the first setup chunk of code we set the timezone to UTC. This is the
time zone at °0 longitude and does not use daylight saving time so it is
always the same. As a default it is good to set this at the beginning of
your script.

```{r timezones}
Sys.setenv(tz="UTC") #set environment time zone to utc

# To view the timezone

Sys.timezone()

# You can set a differnt timezone, e.g.

Sys.setenv(tz="CET") #set environment time zone to central european time
Sys.timezone()

# Let's set it back to UTC for now

Sys.setenv(tz="UTC") 
Sys.timezone()

# We can also get R to tell us the current date and time

Sys.Date()
Sys.time()

# We can do this with the lubridate package too

lubridate::today()
lubridate::now()

# We can even get it to tell us the time in a different zone, say if I want to know if it's a good time to call my kiwi relatives.

lubridate::now(tzone = "NZ")

```

OlsonNames() will show you all the timezone names.

#### Daylight savings time

One of the complications of time is that depending on the time zone and
whether the data has been recorded in local time, it may inconveniently
shift by an hour for part of the year. For example if your data was
collected in the UK in the summer time it may be in British Summer Time,
which is UTC+1. This gets further complicated if you have data that
spans the clock change. [timeanddate.com](timeanddate.com) gives
information about when daylight savings is used in each timezone e.g.
<https://www.timeanddate.com/time/zones/cet> But I would recommend to
always use UTC as your sys timezone! If you want a bit more info about
timezones read this [R blog about timezone
troubles](https://www.r-bloggers.com/2018/07/a-tour-of-timezones-troubles-in-r/)

Now let's learn how to format dates and times in R

#### Formatting dates and times

Your data may store this information in many different ways. Maybe you
have years, days and months in separate columns, maybe months are in
numbers, maybe in words, maybe in 3 letter abbreviations. Maybe your
dates are in UK format, maybe American style with the day and month
switched. Maybe your datetime is all in one column, maybe date and time
are separate. Perhaps you have a separate column that tells you the
timezone (if so, hallelujah someone thought to include this vital
info!). Times may be in 12 or 24 hour format, and may or may not include
seconds Here are some examples:

08/01/2024

2024/08/01

01/08/24

8th Jan 2024

08/01/2024 14:03

08/01/2024 02:03 PM

8 Jan 2024 14 03 06

08/01/2024 15:03 CET

202401081T140306Z

1704722580

You can probably read and understand all of those (except the last one
probably), but R needs to be told what they all are so it knows what
datetime to associate with your data.

Let's take a look at datetime classes. The main ones used are Date (a
day stored as the number of days since 1970-01-01), and POSIXct (date
and time in seconds with the number of seconds beginning at 1 January
1970)

```{r date class}
class(Sys.Date()) # Date class

class(lubridate::now()) #POSIXct class


```

Now let's take some of the date formats above and tell R how to parse
them

```{r date parsing}
date <- "2024/08/01"
str(date)

# We see that R just has this stored as a string, we need to tell R it is a date.

date1 <- as.Date(date)

str(date1) # now stored as a date

#However, if we had give it a date like this:

date <- "01/08/2024"
date2 <- as.Date(date)
str(date2)
# Yyou can see something has gone wrong, it has confused the year and the day. This is because the format wasn't what it expected, so we need to specify the format.

date2 <- as.Date(date, format = "%d/%m/%Y")
str(date2)
# Now it is correct as we told it what to expect.

# as.Date is a base R function, we can also use the very similar lubridate::as_date function

```

Exercise - can you take some of the date formats above and parse them so
R understands them? Use [this
guide](https://www.statology.org/r-date-format/) to understand how to
specify the format.

```{r date exercise}

date1 <- "08/01/2024"

date2 <- "2024/08/01"

date3 <- "01/08/24"

date4 <- "8th Jan 2024"

```

What about when we also have times? Let's look at how to format these

```{r datetime parsing}
datetime <- "08/01/2024 14:03:00"

# Parse this to posixct, specifying the format

datetime1 <- as.POSIXct(datetime, format = "%d/%m/%Y %H:%M:%S", tz="UTC")

# Note we specify the timezone, always a good idea! But if you don't it will assume what you set as your sys.timezone

str(datetime1)

# What about this format?

datetime <- "20240108T140306Z"
datetime2 <- as.POSIXct(datetime, format = "%Y%m%dT%H%M%SZ", tz="UTC")
str(datetime2)

# We can also see how this is as a number
as.numeric(datetime2)

# This should help you with the last format "1704722586" which is actually the simplest as it is the number of seconds since 1st Jan 1970. 
datetime2 <- as.POSIXct(1704722586, origin = "1970-01-01")
str(datetime2)

# 1st Jan 1970 is the standard to count from, but sometimes you get numeric date times with different origins - usually 1st Jan 1900. Just change the origin to parse these.

# Sometimes we have the information spread out, e.g.

Date = c("08/01/2024")
Time = c("14:03:00")

# We can use paste to combine these into one datetime

datetime3 <- as.POSIXct(paste(Date, Time, sep = " "), format = "%d/%m/%Y %H:%M:%S")
str(datetime3)

# If you have your information in a data frame, the parsing is pretty similar, just specify the columns. For example we could create a data frame of the course dates.

course_dates <- data.frame(Date = c("08/01/2024", "09/01/2024", "22/01/2024", "23/01/2024"),
                    StartTime = c("09:00:00", "09:00:00", "09:00:00", "09:00:00"),
                    EndTime = c("16:00:00", "16:00:00", "16:00:00", "15:00:00"),
                    TimeZone = c("UTC", "UTC", "UTC", "UTC"), 
                    Location = "Zicer 0.01", "Zicer 0.01", "Sci 3.05", "EH 1.25")

# Then we can create datetime columns for the start and end times

course_dates$StartDateTime <- as.POSIXct(paste(course_dates$Date, course_dates$StartTime, sep = " "), format = "%d/%m/%Y %H:%M:%S", tz = "UTC")

course_dates$EndDateTime <- as.POSIXct(paste(course_dates$Date, course_dates$EndTime, sep = " "), format = "%d/%m/%Y %H:%M:%S")


```

Exercise - Now we have the start and end date times, can you calculate
how many hours each day is? Hint use the difftime function

```{r duration, error = TRUE, eval=FALSE}

course_dates$duration <- difftime()

```

Once you have your dates parsed you can then extract info from them.

```{r extract date components}
# Extract the year

course_dates$Year <- lubridate::year(course_dates$StartDateTime)
course_dates$Month <- lubridate::month(course_dates$StartDateTime, label = TRUE) #Label = TRUE gives it as month abbr (Jan etc), false gives it as a number

# You can also convert between formats e.g.

course_dates$Month_long <- month.name[course_dates$Month]

# You can also use the format function to extract components

course_dates$day <- format(course_dates$StartDateTime,"%d")
```

## Exercise

Read in the example dataset and create a posixct datetime field. Create
a "Year" and "Month" column

```{r exercise datetime}
# This data is part of the environment agencies water quality archive https://environment.data.gov.uk/water-quality/view/download
ea <- read.csv("./data/ENS-2023.csv")
ea$sample.sampleDateTime[1:10]
ea$datetime <- ""
ea$Year <- ""
ea$Month <- ""
```



### Exercise:

Now it's over to you! Have a look at your own data, or alternatively use
the `mtcars` dataset to practice what you have learned in this session.
Some prompts to think about when exploring and wrangling your data:

-   Are your data in tidy format?

-   Can you select a single column?

-   Can you filter the data?

-   Can you summarise your data?
