---
title: "r_functions_loops"
output: html_document
date: "2025-06-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions

An example of a function in excel might be =SUM(A1:A5), this is code! in R they work the same way, except we can make custom functions to suit our needs.

We use functions in R to make code reusable, so we don’t have to rewrite the same logic multiple times.

They help organize code into clear, manageable sections, making it easier to read and debug.

Functions also allow us to test and reuse specific tasks—like calculating averages or filtering data—across different projects.

The syntax of a function in R is:

```{r function synatx}

function_name <- function(arguments) {
  # code to execute
  return(value) # optional
}

```

#### This function takes a single argument 'name' and returns a greeting message.

```{r basic function}

# Basic function structure
greet <- function(name) {
  
  paste("Hello,", name)
  # the paste commmand concatenates strings together

}


# Try it
greet("James")

greet("Ben")

```

#### We can also supply two arguments to a function, which allows us to perform operations that require more than one input.

```{r double argument function}

# Function with multiple arguments
multiplytwonumbers <- function(a, b) {
  
  product <- a * b
  
  # the return command is used to return a value from the function
  return(product)
  
}

# Try it
multiplytwonumbers(3, 4)

# Try it
multiplytwonumbers(300, 4450)

```

#### Finally, we might want two outputs from a function, such as both the sum and product of two numbers.

Below we can see how to create a function that returns multiple values using a list.

```{r double argument function with return}

# Function with multiple arguments
add_multiply <- function(a, b) {
  
  sum <- a + b
  product <- a * b
  return(list(sum = sum, product = product))
}

# Try it
add_multiply(3, 4)

```

Functions can get extremely complex, but the basic structure remains the same. We define a function with `function_name <- function(arguments) { ... }`, and we can use `return()` to specify what the function should output. So not matter what happens in the middle, we can always return a value OR an error.

Error handling is also an important part of functions. We can use `stop()` to throw an error if something goes wrong, or `warning()` to give a warning without stopping execution.

Lsets trying making our own function.

```{r make your own function}

# Create your own function here

# Example: Function to calculate the square of a number
# Or do whatever

```

But we don't have time to cover that today, so let's move on to loops.

# Loops

Loops are used to repeat (or iterate) a block of code multiple times, which is useful when we want to perform the same operation on different data points or when we need to iterate through a sequence.

We might use for loops to iterate over a vector or list, or while loops to repeat a block of code until a certain condition is met.

We can iterate over anything, it could be a vector, a list, or even a data frame, or even a list of files!

This is a NIGHTMARE to do in excel.

### For Loops

A for loop is used to iterate over a sequence (like a vector or list) and perform an operation on each element.

```{r for loop}

# For loop example
for (i in 1:5) {
  print(paste("Iteration:", i))
}

```

Notice the syntax: `for (variable in sequence) { ... }`. The loop will run for each element in the sequence, assigning the current element to the variable.

```{r loops}


# lets say we want to add 1 to each number in a vector
for (i in 1:5) {
  
  result <- i + 1
  
  print(result)
  
  # we can add as much code as we want inside the loop

}

```
 
The above was on a list of numbers, perhaps we want to iterate over a vector of names instead.

```{r for loop with names}

# For loop with a vector
my_vector <- c("apple", "banana", "cherry")

for (fruit in my_vector) {
  print(paste("Fruit:", fruit))
}

```

or finally a dataframe.

```{r for loop with data frame}

# For loop with a data frame
my_data <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35)
)

for (i in 1:nrow(my_data)) { # nrow(my_data) gives the number of rows in the data frame
  print(paste("Name:", my_data$Name[i], "- Age:", my_data$Age[i]))
  # print(paste("Name:", my_data[i, "Name"], "- Age:", my_data[i, "Age"]))
}


```

### While Loops

These exist in R but might be more frequently used in other languages. They repeat a block of code as long as a specified condition is true. 

```{r while loop}

# While loop example  
counter <- 1

while (counter <= 5) {
  print(paste("Counter:", counter))
  counter <- counter + 1 # Increment the counter
}


```
# Using a funtion inside a loop!

It is often that we might want to perform multiple functions (built in or otherwise) on a single object to transfor it as we go. For example we might start with a number, want to multiple it, then add it to the end of a string. 

But we might have a sequence of numbers to do that too. 

```{r function in loop}

# Function to process a number
# Here we will double it

process_number <- function(x) {
  
  return(x * 2)
  
}

# Now we need a function to add the " is the result" to the end of a string
add_suffix <- function(x) {
  
  return(paste(x, "is the result"))
  
}

# Then we can use both functions in a loop, iterating over numbers
numbers <- c(1, 2, 3, 4, 5)

for (num in numbers) {
  
  processed_num <- process_number(num) # Process the number
  
  # Now we can add the suffix to the processed number
  result_string <- add_suffix(processed_num) # Add suffix to the result
  
  print(result_string) # Print the final result
  
}

```

The above is much like a normal workflow in R. As you have seen earlier and tomorrow. 








