---
title: "Day1_BestCoding"
author: "Ellie Harris"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What are best coding practices?

We use best coding practices to help make our R code easier to **read**, **write**, **share**, and **verify**. The [Google style guide](https://google.github.io/styleguide/Rguide.html), and [Tidyverse Style guide](https://style.tidyverse.org/) are a good place to start for general rules of thumb and common preferences. The tidyverse guide was created from Google's initial style guide, whilst Google's new style guide is based on the tidyverse style guide, so they are very similar!

## File names

File names should be meaningful and end in `.R`. Avoid using special characters in file names - stick with numbers, letters, `-`, and `_`.

``` {#FileNames1}
# Good
fit_models.R
utility_functions.R

# Bad
fit models.R
Random.r
stuff.r
```

If files should be run in a particular order, prefix them with numbers. If it seems likely you'll have more than 10 files, left pad with zero:

``` {#FileNames2}
00_download.R
01_explore.R
...
09_model.R
10_visualize.R
```

# Key structural points

## About

When you look back at your code, or send it to others, you want to be able to navigate it easily, understand what you've done, and easily reproduce it.

What is the R script for? And who wrote it?

You can add an 'About' section to the top of the script. It can be in any format but should contain the author, purpose of the script, and when it was made

```{r About, eval=FALSE, message=FALSE, warning=FALSE}

# Script to outline best coding practices in R
# Author: Ellie Harris
# Last update: 05.06.2025
```

## Set-up

At the start of the script, explicitly state the packages used. It's useful to have a set-up and/or data section where everything needed is loaded at the start. You can also label the use of the package, which is especially helpful for one which may only be used for one small element in the script.

```{r Setup, eval=FALSE, message=FALSE, warning=FALSE}

## Setup -----------------------------------------------------------------------

library(terra) # for spatial analyses
library(data.table) # for data tables
library(paletteer) # for colour palette
```

## Navigation

Scripts can get long... really long. They're also usually comprised of multiple sections. To navigate easily through the script, it's good practice to use headings and sub headings.

```{r Navigation, eval=FALSE, message=FALSE, warning=FALSE}
# Data --------------------------------------------------------------------------
## Climate ----

## Species occurrence ----

```

Using multiple '\#' will make subheadings and subsubheadings, which can be viewed on the 'Outline tab' at the top right of the script.

## Data

Read in data and directories early on so they can be referenced directly rather than continuously read in.

Ensure your working directory is set before reading in files, preferably through an Rproject or through the 'here' package rather than 'setwd()'.

```{r Data files, eval=FALSE, message=FALSE, warning=FALSE}

# Read in data
input_file <- "data/data.csv" 
output_file <- "data/results.csv"

output_dir <- "./Output/"
```

## Comments

Each line of a comment should begin with the comment symbol and a single space: `#`

Comments should be used to explain the 'why' and not the 'what' or 'how', but use them as makes sense for you and in whatever way makes your code clearer and more easy to follow.

If you discover that you have more comments than code, consider switching to [R Markdown](https://rmarkdown.rstudio.com/).

Comments should be in sentence case, and only end with a full stop if they contain at least two sentences

```{r, eval = FALSE}
# Good

# Objects like data frames are treated as leaves
x <- map_if(x, is_bare_list, recurse)

# Do not use `is.list()`. Objects like data frames must be treated
# as leaves.
x <- map_if(x, is_bare_list, recurse)


# Bad

# objects like data frames are treated as leaves
x <- map_if(x, is_bare_list, recurse)

# Objects like data frames are treated as leaves.
x <- map_if(x, is_bare_list, recurse)
```

## Keep clean

Keep a clean working environment. Just like you might keep your desk tidy, a cluttered R environment should be kept clean to avoid mistakes and unnecessary memory usage

```{r cleanup, eval=FALSE, message=FALSE, warning=FALSE}

# Sample dataset of 1000 rows
dat <- data.frame(A = 1:10, B = 1:10, C = 21:30)
object.size(dat) # Reports the memory size allocated to the object

# Remove object from your environment... BE CAREFUL!!
rm("dat") 

# Release unused memory
gc() 

# List all objects in your environment
ls() 

# Delete all objects in your workspace - BE CAREFULL! R doesn't ask is you're sure, and you can't recover them. I'd recommend only putting this at the end of a script after saving all necessary data
rm(list = ls()) 
```

# Within-script details

## Assignment

Use `<-`, not `=`, for assignment.

```{r}
# Good
x <- 5

# Bad
x = 5
```

## Object names

Variable and function names should use only lowercase letters, numbers, and `_`. Use underscores (`_`) to separate words within a name.

```{r, eval = FALSE}
# Good
day_one
day_1

# Bad
DayOne
dayone
```

Where possible, avoid re-using names of common functions and variables. This will cause confusion for the readers of your code.

```{r, eval = FALSE}
# Bad
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```

## Spacing

Always put a space after a comma, never before, just like in regular English.

```{r, eval = FALSE}
# Good
x[, 1]

# Bad
x[,1]
x[ ,1]
x[ , 1]
```

Do not put spaces inside or outside parentheses for regular function calls.

```{r, eval = FALSE}
# Good
mean(x, na.rm = TRUE)

# Bad
mean (x, na.rm = TRUE)
mean( x, na.rm = TRUE )
```

Place a space before and after `()` when used with `if`, `for`, or `while`.

```{r, eval = FALSE}
# Good
if (debug) {
  show(x)
}

# Bad
if(debug){
  show(x)
}
```

Place a space after `()` used for function arguments:

```{r, eval = FALSE}
# Good
function(x) {}

# Bad
function (x) {}
function(x){}
```

### Infix operators

Most infix operators (`==`, `+`, `-`, `<-`, etc.) should always be surrounded by spaces:

```{r, eval = FALSE}
# Good
height <- (feet * 12) + inches
mean(x, na.rm = TRUE)

# Bad
height<-feet*12+inches
mean(x, na.rm=TRUE)
```

There are a few exceptions, which should never be surrounded by spaces:

-   The operators with [high precedence](https://rdrr.io/r/base/Syntax.html): `::`, `:::`, `$`, `@`, `[`, `[[`, `^`, unary `-`, unary `+`, and `:`.

    ```{r, eval = FALSE}
    # Good
    sqrt(x^2 + y^2)
    df$z
    x <- 1:10

    # Bad
    sqrt(x ^ 2 + y ^ 2)
    df $ z
    x <- 1 : 10
    ```

-   The help operator

    ```{r, eval = FALSE}
    # Good
    package?stats
    ?mean

    # Bad
    package ? stats
    ? mean
    ```

## Function calls

### Named arguments

A function's arguments typically fall into two broad categories: one supplies the **data** to compute on; the other controls the **details** of computation. When you call a function, you typically omit the names of data arguments, because they are used so commonly. If you override the default value of an argument, use the full name:

```{r, eval = FALSE}
# Good
mean(1:10, na.rm = TRUE)

# Bad
mean(x = 1:10, , FALSE)
mean(, TRUE, x = c(1:10, NA))
```

## Long lines

Strive to limit your code to 80 characters per line. This fits comfortably on a printed page with a reasonably sized font. If you find yourself running out of room, this is a good indication that you should encapsulate some of the work in a separate function.

If a function call is too long to fit on a single line, use one line each for the function name, each argument, and the closing `)`. This makes the code easier to read and to change later.

```{r, eval = FALSE}
# Good
do_something_very_complicated(
  something = "that",
  requires = many,
  arguments = "some of which may be long"
)

# Bad
do_something_very_complicated("that", requires, many, arguments,
                              "some of which may be long"
                              )
```

As described under [Named arguments](#argument-names), you can omit the argument names for very common arguments (i.e. for arguments that are used in almost every invocation of the function). Short unnamed arguments can also go on the same line as the function name, even if the whole function call spans multiple lines.

```{r, eval = FALSE}
map(x, f,
  extra_argument_a = 10,
  extra_argument_b = c(1, 43, 390, 210209)
)
```

You may also place several arguments on the same line if they are closely related to each other, e.g., strings in calls to `paste()` or `stop()`. When building strings, where possible match one line of code to one line of output.

```{r, eval = FALSE}
# Good
paste0(
  "Requirement: ", requires, "\n",
  "Result: ", result, "\n"
)

# Bad
paste0(
  "Requirement: ", requires,
  "\n", "Result: ",
  result, "\n")
```

## Semicolons

Don't put `;` at the end of a line, and don't use `;` to put multiple commands on one line.

## Data

### Character vectors

Use `"`, not `'`, for quoting text. The only exception is when the text already contains double quotes and no single quotes.

```{r, eval=FALSE}
# Good
"Text"
'Text with "quotes"'
'<a href="http://style.tidyverse.org">A link</a>'

# Bad
'Text'
'Text with "double" and \'single\' quotes'
```

### Logical vectors

Prefer `TRUE` and `FALSE` over `T` and `F`.

## Automatic tidying

In R markdown, code can be automatically tidied using 'tidy=TRUE' in the code chunk header.

```{r eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
# messy R code...
1+1
x=1:10
if(TRUE){
  
print('Hello world!'  )}

ans=sum(  5+    5) 

```

The clean version:

```{r, tidy=TRUE, echo = TRUE, eval = FALSE, include = TRUE,  tidy.opts=list(arrow=TRUE, indent=2)}

#{r, tidy=TRUE, echo = TRUE, eval = FALSE, include = TRUE,  tidy.opts=list(arrow=TRUE, indent=2)}

# messy R code…
1+1
x=1:10
if(TRUE){
  
print('Hello world!'  )}

ans=sum(  5+    5) 
```

------------------------------------------------------------------------

# Finally...

Here are some useful packages

-   [styler](http://styler.r-lib.org) allows you to interactively restyle selected text, files, or entire projects. It includes an RStudio add-in, the easiest way to re-style existing code.

-   [lintr](https://github.com/jimhester/lintr) performs automated checks to confirm that you conform to the style guide.

------------------------------------------------------------------------

# For later:

## Functions

### Naming

As well as following the general advice for [object names], strive to use verbs for function names:

```{r eval = FALSE}
# Good
add_row()
permute()

# Bad
row_adder()
permutation()
```

### Long lines

There are two options if the function name and definition can't fit on a single line:

-   Function-indent: place each argument on its own line, and indent to match the opening `(` of `function`:

    ```{r, eval = FALSE}
    long_function_name <- function(a = "a long argument",
                                   b = "another argument",
                                   c = "another long argument") {
      # As usual code is indented by two spaces.
    }
    ```

-   Double-indent: Place each argument of its own **double** indented line.

    ```{r, eval = FALSE}
    long_function_name <- function(
        a = "a long argument",
        b = "another argument",
        c = "another long argument") {
      # As usual code is indented by two spaces.
    }
    ```

In both cases the trailing `)` and leading `{` should go on the same line as the last argument.

Prefer function-indent style to double-indent style when it fits.

These styles are designed to clearly separate the function definition from its body.

```{r, eval = FALSE}
# Bad
long_function_name <- function(a = "a long argument",
  b = "another argument",
  c = "another long argument") {
  # Here it's hard to spot where the definition ends and the
  # code begins, and to see all three function arguments
}
```

If a function argument can't fit on a single line, this is a sign you should rework the argument to keep it [short and sweet](https://design.tidyverse.org/def-short.html).

### `return()`

Only use `return()` for early returns. Otherwise, rely on R to return the result of the last evaluated expression.

```{r eval = FALSE}
# Good
find_abs <- function(x) {
  if (x > 0) {
    return(x)
  }
  x * -1
}
add_two <- function(x, y) {
  x + y
}

# Bad
add_two <- function(x, y) {
  return(x + y)
}
```

Return statements should always be on their own line because they have important effects on the control flow. See also [inline statements](#inline-statements).

```{r, eval = FALSE}
# Good
find_abs <- function(x) {
  if (x > 0) {
    return(x)
  }
  x * -1
}

# Bad
find_abs <- function(x) {
  if (x > 0) return(x)
  x * -1
}
```

If your function is called primarily for its side-effects (like printing, plotting, or saving to disk), it should return the first argument invisibly. This makes it possible to use the function as part of a pipe. `print` methods should usually do this, like this example from [httr](http://httr.r-lib.org/):

```{r eval = FALSE}
print.url <- function(x, ...) {
  cat("Url: ", build_url(x), "\n", sep = "")
  invisible(x)
}
```

## If statements

-   If used, `else` should be on the same line as `}`.

-   `&` and `|` should never be used inside of an `if` clause because they can return vectors. Always use `&&` and `||` instead.

-   NB: `ifelse(x, a, b)` is not a drop-in replacement for `if (x) a else b`. `ifelse()` is vectorised (i.e. if `length(x) > 1`, then `a` and `b` will be recycled to match) and it is eager (i.e. both `a` and `b` will always be evaluated).

    If you want to rewrite a simple but lengthy `if` block:

    ```{r, eval = FALSE}
    if (x > 10) {
      message <- "big"
    } else {
      message <- "small"
    }
    ```

    Just write it all on one line:

    ```{r, eval = FALSE}
    message <- if (x > 10) "big" else "small"
    ```

## 

## Pipes

### Introduction

Use `%>%` (tidyverse, covered later) or \|\> (base R) to emphasise a sequence of actions, rather than the object that the actions are being performed on.

Avoid using the pipe when:

-   You need to manipulate more than one object at a time. Reserve pipes for a sequence of steps applied to one primary object.

-   There are meaningful intermediate objects that could be given informative names.

This will be covered in more detail in the tidyverse section of the course, but here are a few small notes on the topic:

### Long lines

If the arguments to a function don't all fit on one line, put each argument on its own line and indent:

```{r, eval = FALSE}
iris %>%
  group_by(Species) %>%
  summarise(
    Sepal.Length = mean(Sepal.Length),
    Sepal.Width = mean(Sepal.Width),
    Species = n_distinct(Species)
  )
```

```{}
```
