---
title: "Data formats"
author: "Sam Speak"
date: "2023-12-12"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data formats

In this session we will explore the different types of data we might come across. How they are formatted in R, as well as how we can import and export our own data.

## Types of Data

We can check what type of data class a variable is using the command `class()`

```{r check class}
#class()
```

### Numeric

In this format values are read as numeric values these can be whole numbers or decimals (floats). i.e 10, 22.5, 10000, π

```{r Numeric}
n <- pi

n

class(n)
```

### Integers

Integers are numeric data formed of whole numbers, without any decimals. You should use this class of data when you are sure that you will not create a variable with any decimals.

To create this type of data you must follow the value with a letter `L`.

```{r Interger}
I <- 30

I

class(I)

i <- 30L

i

class(i)

```

#### Some quick maths

This means that we can use R for maths. We can add `+`, subtract `-`, times `*` and divide `/`.

```{r Maths}

#### We can make numeric variables by adding numbers together
n <- 3 + 4

#### Or we can take them away from each other 
m <- 7 - 4

#### This can be done with integers (remember to put the L after the numbers)
u <- 30L - 3L

u

class(u)
```

```{r multiplication}
#### We can multiply variables together
x <- n * m

x

class(x)

#### We can divide variables by numbers
y <- x/2

y

class(y)
```

```{r}
#### Or by other numeric variables 
π <- pi 

t <- π/y 

t

class(t)

#### We can also raise to the power 

k <- 2^3

k

class(k)

```

### Task 1

### Try this for yourself:

```{r Maths task}

#{type your own maths here}

```

### Characters/Strings

Text that is kept as text it should be inputted with `""` or `''` around the values. Either work but the opening and closing quotes need to match.

i.e "Hello World" == 'Hello World'

```{r Character}
ch <- "R is my favourite language"

ch

class(ch)
```

Characters can be combined together

```{r Character merge}
part_a <- "What is your favourite food?"

part_b <- 'Mine would have to be Lasagne!' 
  
whole <- paste(part_a, part_b)

print(whole)

class(whole)

```

We can use paste to add in text between values. You could use this to add in underscores `_` or other text between character strings.

`paste` puts strings together with a space between each (or whatever you specify with the "sep" argument), `paste0` leaves no gap between the strings.

```{r paste}

part_a

part_b

longer_whole <- paste(part_a, "But if you asked me .... ", part_b)

longer_whole

sequence_a <- "AGTCGATCGATGCTAGAA"

sequence_b <- "CGTAGCTCAGTCGATC"

longer_sequence <- paste0(sequence_a, sequence_b)
longer_sequence

longer_sequence <- paste(sequence_a, sequence_b)
longer_sequence

longer_sequence <- paste(sequence_a, sequence_b, sep = "_")
longer_sequence



```

Characters can be combined together

We can check the length of characters using the function `nchar()`

```{r Character length}

book_1 <- "Last night I dreamt I went to Manderley again. It seemed to me I stood by the iron gate to the drive, and for a while I could not enter"

nchar(book_1)

seq_1 <- "AACCGTAGCTCAGTCGATCGATGCTAGAACTAGCCATG"

nchar(seq_1)

```

#### Escape characters

If we wanted to add quotes to a character string we find a problem

```{r Escape Character problem, eval=F}

Roosevelt <- "It is always good to learn new things we should "Do one thing every day that scares you. " like learning R"  

```

This gives us an error as the quotation marks break up the sentence, so R thinks the phrase should end after the first quote and is confused about what comes after. So we need to use the escape character `\` when adding quotation marks.

```{r Escape Character}

Roosevelt <- "It is always good to learn new things we should \"Do one thing every day that scares you.\" like learning R" 

Roosevelt
```

There are multiple escape characters that we can use when inputting text.

| R code | Result          |
|--------|-----------------|
| \\"    | Quotation marks |
| \\\\   | Backslash       |
| \\n    | New line        |
| \\r    | Return carriage |
| \\t    | Tab             |
| \\b    | Backspace       |

### Logical/Boolean

We often need to check if values are `TRUE` or `FALSE` these are known as a Boolean or Logical value.

```{r Logical}
l <- TRUE

l

class(l)
```

You can compare two values, R will then inform you of the logical answer.

```{r Logical questions}

100 > 10 

100 == 10

100 < 10

#### You can also save the result of a question as a bool value 
a <- 20
# Is 100 greater than a?
q <- 100 > a

q

##### These values do not automatically update KC - do not understand what sam meant here
a <- 400
q <- 100 > a

q

class(q)

```

### Task 2

Try it for yourself here:

Make a variable of each class: Numeric, Integer, Character and Logical.

```{r Task 2}




```

## Data Structures

### Vectors

So far we have been working with just items. But there are multiple different data formats that we can work with.

The first kind is Vectors, this is a means of combining a list of items that are of the same type of variable.

These vectors are created using the `c()` function and the items need to be comma separated.

#### Creating vectors

```{r making vectors}
## Vectors can be Strings 
films <- c("Barbie", "Oppenheimer", "Paddington", "Wonka", "Dune")

films

## Numerical 

numbers <- c(1, 3, 5, 7, 9)

numbers

## Booleans

Bools <- c(TRUE, TRUE, FALSE, TRUE, FALSE, FALSE)

Bools

# We can examine the structure of these

str(films)
str(numbers)
str(Bools)

```

#### Length

Unlike variables we can evaluate each item in a vector independently this means we check the length of vectors with `length()`. This will tell us how many items the vector has.

```{r length vectors}

length(numbers)

length(Bools)

```

#### Sort

This allows us to sort the values within a vector

```{r sort vectors}

sort(numbers)

#### We can sort non numerical values as well

#### How are strings sorted?
sort(films)

#### what about bools?
sort(Bools)

```

#### Accessing values

Vectors allow us to access specific values using the `[]`.

```{r accessing vectors}

films[2] # access the 2nd film

numbers[5] # access the 5th number

Bools[3] # access the 3rd boolean

```

We can access not only values but multiple values with the same comma function `c()`

```{r multiple values}

films[c(1,4)] # access the 1st and 4th film

```

#### Changing values

```{r changing vlaues}

films 

### We can add values to vectors 
films[6] <- c("Maverick")

## We can also change values that are already in a vector
films[3] <- c("Paddington 2") 

films

```

#### Sequences

We can also make vectors of a sequence of numeric values using `:`

```{r sequence vectors}
number_sequence <- 1:20

number_sequence

## These can also be decimals
decimals <- 1.5:8.5

decimals

decimals <- 1.5:7.8

decimals

```

We can add steps to this sequence using the function `seq()`

```{r sequence steps}
number_sequence_2 <- seq(from = 0, to = 200, by = 50)

number_sequence_2

```

### Task 3

1.  How many items are in the Vector films & what is the longest film name?

2.  How many values between 0 and 300 are dividable by 4 with no remainder?

```{r task 3}
# 1)

# 2)

```

### Lists

Lists are similar to vectors but can contain data of different types.

They are created by using the function `list()`

```{r lists}

list_fruit <- list("apple", "pear", "banana", "coffee")

list_fruit

numbers <- list(1, "three", 5, 7, 9)

str(numbers)

```

We can access values in a list in the same way as vectors using `[]`

```{r accessing values}

list_fruit[2:3]

#sort(list_fruit)

length(list_fruit)

```

#### Adding to lists

We can add values to lists in two different ways

```{r adding to lists}

list_fruit[5] <- c("cherry")

list_fruit <- append(list_fruit, values = "Mango") 

```

#### Checking values

We can also check to see if values are in a list using `%in%`

```{r checking values}

"raspberry" %in% list_fruit
"cherry" %in% list_fruit

```

### Task 4

1.  Make your own list of your favorite 3 foods

1b. Make a list of odd numbers between 1 and 15

1c. Combine these two lists

1d. Add in a new item to be second in each list

```{r Task 4}



```

### Data frames

A data frame is a format of data displayed as a table

Each column in a data frame can be a different format but data within columns should be in the same format.

To create a data frame we use the function `data.frame()`

```{r making df}

Animals_df <- data.frame(
  Animal = c("Dog", "Cat", "Shark", "Whale", "Bear"),
  Length.m= c(0.8, 0.5, 2.1, 15, 2.4),
  Age.years = c(11, 12, 21, 34, 18)
)

Animals_df

```

#### Accessing data

There are 3 ways to access data in a data frame.

Numerically as with vectors and lists `[]`

Naming the column we want `[[]]`

or the easiest with `$` (press `Tab` and you can list all the columns to select)

```{r acessing df}

Animals_df[2]

Animals_df[["Animal"]]

Animals_df$Age.years
```

#### Adding data

We can add data to our data frames in a variety of ways. We can use the functions `rbind()` and `cbind()` to join two data frames together by their rows or columns respectively.

```{r adding data to dataframes}

Animals_df[["Diet"]] <- c("Omnivore", "Carnivore", "Pescitarian", "Krill", "Omnivore")

Animals_df

Animals_df_2 <- rbind(Animals_df, c("Cow", 3, 10, "Herbivore"))

Animals_df_2

Animals_df_3 <- cbind(
              Animals_df_2, 
              Habitat = c("Sofa", "House", "Marine", "Marine", "Arboreal", "Grassland"),
              common_name = c("Dog","Cat","Shark","Whale","Bear","Cow"),
              latin_name = c("Canis lupus familiaris", "Felis catus", "Carcharhinus longimanus", "Megaptera novaeangliae", "Ursus arctos", "Bos taurus")
              )

Animals_df_3


```

### Task 5

Add a new animal to the data frame using `rbind()`

Add the speed of each animal using `cbind()`

```{r Task 5}


```

#### Size of Data frames

We can find the dimensions of a data frame with `dim()`. As well as `ncol()` and `nrow()`

```{r dimensions of dataframes}

dim(Animals_df_3)

ncol(Animals_df_3)

nrow(Animals_df_3)

```

We can also find out the names of the columns or rows with `colnames()` and `rownames()`

```{r names in dataframes}

colnames(Animals_df_3)

rownames(Animals_df_3)

### We can change these too 

colnames(Animals_df_3)

colnames(Animals_df_3) <- c("Animal",
                           "Length.m",
                           "Age.years",
                           "Diet",
                           "Habitat",
                           "Common_name",
                           "Latin_name")

colnames(Animals_df_3)

```

### Matrix

A matrix (a 2D) can be created with the `matrix()` function. Specify the `nrow` and `ncol` parameters to get the amount of rows and columns. They only have two dimensions rows and columns.

```{r matrix }

my_matrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3, ncol = 2)

my_matrix

### We can also create matrix of strings too

another_matrix <- matrix(list_fruit, nrow = 2, ncol = 3)

another_matrix

```

You can access the items by using `[]` brackets. The first number in the bracket specifies the row-position, while the second number specifies the column-position

```{r acessing matrix}

my_matrix[3, 2]

#### If we leave either value empty we get the whole row or column 

my_matrix[, 2]

my_matrix[1, ]

### We can also access more than one column or row

another_matrix[, c(1, 2)]

```

#### Adding to a matrix

We can add to a matrix using `cbind()` and `rbind()`. Make sure to include the `c()`

```{r adding matrix values}

fruit_matrix <- rbind(another_matrix,c("orange", "pineapple", "grape"))

fruit_matrix

```

We can remove items using `-c()` and the row and column numbers to remove

```{r removing matrix values}

shopping_list <- fruit_matrix[-c(1), -c(1)]

shopping_list

#### What will happen if we only have one value 

fruit_matrix[-c(2), ]

```

We can check values like we do for lists with `%in%`

```{r checking matrix values}

"apple" %in% fruit_matrix

"apple" %in% shopping_list

```

### Arrays

Arrays are similar to Matrices but they can have more than two dimensions.

```{r making an array}
my_first_array <- c(1:64)
my_first_array

# An array with more than two dimensions
my_second_array <- array(my_first_array, dim = c(8, 4, 2))
my_second_array

```

We can access the array elements by referring to the index position. We can use the `[]` brackets to access the desired elements from an array.

```{r accessing array}
my_first_array <- c(1:64)
my_second_array <- array(my_first_array, dim = c(8, 4, 2))

my_second_array[2, 3, 2]

```

The syntax is as follow: array[row position, column position, matrix level]

You can also access the whole row or column from a matrix in an array, by using the `c()` function.

A comma `,` before `c()` means that we want to access the column.

A comma `,` after `c()` means that we want to access the row.

```{r array multiple values}
my_second_array <- array(my_first_array, dim = c(8, 4, 2))
my_second_array[c(4), , 1]

# Access all the items from the first column from matrix one
my_second_array <- array(my_first_array, dim = c(8, 4, 2))
my_second_array[,c(3), 2]
```

### Factors

Factors are used to categorize data and have levels. Examples of factors are:

Music: Rock, Pop, Classic, Jazz

Training: Strength, Stamina

Organisation: ARIES, CEFAS

To create a factor, use the `factor()` function and add a vector as argument.

```{r making factors}

film_types <- factor(
  c("Comedy", "Action", "Horror", "Romance", "Action", "Horror", "Comedy", "Comedy")
  )

film_types

```

#### Levels

We can check the levels that a factor has with the command `levels()`

We can also use this to define the levels in a factor when we make it.

```{r factors levels}

levels(film_types)

film_types <- factor(
  c("Comedy", "Action", "Horror", "Romance", "Action", "Horror", "Comedy", "Comedy"),
  levels = c("Action", "Comedy", "Horror", "Romance")
  )

film_types
```

#### Changing factor values

We can access factors using the `[]` brackets. This means we can also change values as we have done before.

```{r changing factors values}

film_types[3]

film_types[4] <- c("Comedy")

film_types

##### Note how "Romance" is still a level of our factor even though none of the values contain "Romance" anymore 

##### We cant add values that are not one of the levels 

film_types[1] <- c("SciFi")

#### We need to first add that as a level


film_types <- factor(
  c("Comedy", "Action", "Horror", "Romance", "Action", "Horror", "Comedy", "Comedy"),
  levels = c("Action", "Comedy", "Horror", "Romance", "SciFi")
  )

film_types[1] <- c("SciFi")

film_types

##### We can edit the levels themselves 

levels(film_types)[6] <- c("Western") 

length(film_types)

film_types[9] <- c("Western")

film_types

```

## Changing data format

We can convert between data formats.

We can do this for whole data frames or just columns.

```{r Changing formats}

str(Animal_Df_4)

Animal_Df_4$Length.m <- as.numeric(Animal_Df_4$Length.m) 

str(Animal_Df_4)

Animal_Df_4$Habitat <- as.factor(Animal_Df_4$Habitat)

str(Animal_Df_4)

Animal_matrix <- as.matrix(Animal_Df_4)

Animal_matrix

#### But this looks a bit messy so lets just change the names

rownames(Animal_matrix) <- Animal_matrix[, 1]

#### And select the columns we need 

Animal_matrix <- Animal_matrix[ , c(2:7)]

Animal_matrix

```

## Creating your own test data

To test out code on what would be a small dataset, you can simply enter test data yourself, e.g.

```{r, echo=TRUE}

df1 <- data.frame(
  ID = c(1, 2, 3, 4, 5),
  var1 = c('a', 'b', 'c', 'd', 'e'),
  var2 = c(1, 1, 0, 0, 1)
  )

df1 
```

You can generate numbers, characters and more with functions.

```{r, echo=TRUE}
seq(1, 16, by=2) # sequence of numbers from 1 to 16 increment by 2.
LETTERS[1:8] # the 8 upper-case letters of the english alphabet.
month.abb[1:8] # the three-letter abbreviations for the first 8 English months
sample(10:20, 8, replace = TRUE) # 8 random numbers with replacement from 10 to 20.
letters[1:8] # the first 8 lower-case letters of the english alphabet.

df2 <- data.frame(
  a = seq(1,16,by=2), 
  b = LETTERS[1:8],
  x = month.abb[1:8],
  y = sample(10:20,8, replace = TRUE),
  z = letters[1:8]
  )

df2

```

Similarly, you can sample from a given set of numbers with or without replacement

```{r, echo=TRUE}
# Generate 15 numbers randomly sampled, with replacement, from the numbers 1, 2, 3
df3 = data.frame(X = sample(1:3, 15, replace = TRUE))

df3
```

You can also generate numbers using mean and standard deviations of sampled data, eg. using rnorm

```{r, echo=TRUE}
set.seed(1) # allows random generations to be replicated
df4 <- data.frame(Y = rnorm(15), Z = ceiling(rnorm(15)))

df4
```

#### Task 6

Create a random data frame 12 rows of 3 variables:

A)  Random numbers between 1 and 50 with replacement

B)  Random numbers between 1 and 20 no replacement

C)  The months of the year

```{r Task 6}

```

## Exploring data

We can see the top few lines of a data frame with `head()`

```{r heads}
mtcars

head(mtcars)

head(mtcars , n = 10)
```

similarly we can see the bottom of the data frame with `tail()`

```{r tail}
tail(mtcars)

tail(mtcars , n = 10)

```

When we load in a data frame we can also see a quick summary of the statistics.

We can see this for every column in the data frame we do this with `summary()`

```{r summary}
summary(mtcars)

```

Or we can see it for just one column

```{r summary col}
summary(mtcars$wt)

```

#### Task 7

How would you view the top 20 lines of mtcars?

Find the mean horse power (hp) of the cars?

```{r Task 7}


```

## Importing data

So far we have only been using data that we have made ourselves. But this is not that common, we often want to load data into R from somewhere else.

We can do this using the `read` function. We can also express what format the data is in.

Read in the chosen file `read.delim()`, you can express the deliminators using `sep` .

Read in a csv (comma separated values) file, `read.csv()` will be expecting the file to be comma separated.

We can also read in Excel files we need to use the `read.excel()` function.

If you have data in a non standard file type it is always a good idea to search for a package to import it.

```{r importing data}

#read.delim("~/PATH/to/your/data/data_file.txt", sep = "/" )
```

Most importantly we can export our data from R to save it as a file. A good example of this is to write out our animal data frame as a csv with `write.csv()`.

```{r exporting data}

##### Please note you will need to put the PATH from where you are to where you want to save the data.

write.csv(Animal_Df_4, file = "~/Documents/Training/Animal_data.csv", row.names = FALSE)

saved_animal_data <- read.csv("~/Documents/Training/Animal_data.csv", header = TRUE)

saved_animal_data

```

Helpful links:

A tutorial on how to import a variety of data formats: <https://www.datacamp.com/tutorial/r-data-import-tutorial>

Some useful packages for reading in different types of data:

`readxl` - for Excel data. <https://www.rdocumentation.org/packages/readxl/versions/0.1.1/topics/read_excel>

`rjson` - Json format files. <https://cran.r-project.org/web/packages/rjson/rjson.pdf>

`RSQLite` - sqlite files. <https://www.rdocumentation.org/packages/RSQLite/versions/2.3.1>

`vcfR` - VCF SNP data into R. <https://cran.r-project.org/web/packages/vcfR/vignettes/intro_to_vcfR.html>

`readPed` - Pedigrees <https://search.r-project.org/CRAN/refmans/pedtools/html/readPed.html>
