---
title: "Welcome to R Studio and R Markdown"
output:
  html_document: default
  pdf_document: default
date: "2024-01-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is R!

# R Studio

-   There is a million ways to do things in R!

-   Fun settings and appearance!

-   Restarting session

    -   Needs to be done at the start of each session

-   Saving and creating new documents

-   Default R data <http://www.sthda.com/english/wiki/r-built-in-data-sets>

## R Markdown VS R Scripts

All training will be delivered using R Markdown, this is so both the instructions and explanation, code and the output can be nicely visualized.

Source OR Visual

R markdown also allows for a document to be produced that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r printhello}

print("Hello welome to our R Course at UEA!")

```

The print bit here is a built in function (its much the same as the functions in excel!)

We use functions to do almost everything... custom functions we will talk about tomorrow.

```{r printmaths}

1+1

```

```{r error}

thiswillcauseanerror() 

```

### Console

?help

[demo]

### Environment

```{r envtext}

sometext <- "happy new year"

number <- 1

```

```{r calltest}

sometext

number + 4

```

```{r dfexample}

dataframe_example <- mtcars

```

### Files, plots, packages and help!

Plots in R Markdown

```{r plot, echo=FALSE}

plot(mtcars$wt, mtcars$mpg)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

Plots in R Scripts

# Packages and modules

[package guide](https://hbctraining.github.io/Intro-to-R-flipped/lessons/04_introR_packages.html)

Packages are collections of R functions, data, and compiled code in a well-defined format, created to add specific functionality. There are 10,000+ user contributed packages and growing.

There are a set of standard (or base) packages which are considered part of the R source code and automatically available as part of your R installation. Base packages contain the basic functions that allow R to work, and enable standard statistical and graphical functions on datasets; for example, all of the functions that we have been using so far in our examples.

However, the more you work with R, you will come to realize that there is a cornucopia of R packages that offer a wide variety of functionality. To use additional packages will require installation. Many packages can be installed from the CRAN or Bioconductor repositories.

```{r findcurrentlibs_first}

sessionInfo()

```

```{r testlib}

ggplot()

```

```{r package install}

install.packages("ggplot2")

library("ggplot2")

library("dynamicSDM")

```

```{r findcurrentlibs}

sessionInfo()

```

```{r loadlib}

ggplot2::

```

Versions in packages - like any software, best to keep them up to date!

# Comments in code

```{r codecomment}

print("hello world")

print("hello world again")

```

# Stitching a document with R Markdown to HTML
