---
title: "functionsandloops"
output: html_document
date: "2025-06-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions

A function in R is a reusable block of code designed to perform a specific task. Instead of writing the same code repeatedly, you can define a function once and call it whenever needed. This not only saves time but also makes your code cleaner, easier to read, and less prone to errors. Functions help break down complex problems into smaller, manageable parts, making your analysis more modular and organized. They are especially useful when working with repetitive tasks, such as data transformations or calculations, and are essential for writing efficient and maintainable R scripts.

Writing custom functions is extremely useful.

It means we can repeat a process again, and again, and again

```{r examplefun}


repeat_again <- function(numberoftimes) {
  rep("and again", times = numberoftimes)
}

# Call the function
repeat_again(50)



say_hello <- function(name) {
  message("Hello, my name is ", name, "!")
}


say_hello("James")

```

Breakdown of the syntax: function_name: This is the name you assign to your function so you can call it later.

`functioname <-` The assignment operator, used to assign the function to the name.

`function(...)` This declares that you're defining a function. Inside the parentheses, you list the arguments (inputs) the function will take. You can also assign default values to arguments.

`{ ... }` The body of the function is enclosed in curly braces. This is where you write the code that performs the desired task.

`return(...)` This is optional but commonly used to specify what value the function should output.

Its functions (and loops) that make code extremly superior to excel , access etc.

You can perform lots of operations in a function

```{r longfun}

long_arithmetic <- function(x) {
  
  x <- x + 5
  x <- x - 3
  x <- x + 12
  x <- x - 7
  x <- x + 20
  x <- x - 15
  x <- x + 8
  x <- x - 4
  x <- x + 10
  x <- x - 6
  x <- x + 14
  x <- x - 9
  x <- x + 11
  x <- x - 2
  x <- x + 7
  x <- x - 5
  x <- x + 13
  x <- x - 8
  x <- x + 6
  x <- x - 1
  
  return(x)
  
}

# Example usage:
long_arithmetic(10)

```

We can add more than one argument to our function

```{r doublearg}


multiply_numbers <- function(a, b) {
  result <- a * b
  return(result)
}

# Example usage:
multiply_numbers(4, 5)



```

As things get more complex, we can add warnings to our functions

```{r warningsfun}


double_number <- function(x) {
  if (!is.numeric(x)) {
    warning("Input must be a numeric value.")
    return(NA)
 }
 return(x * 2)
}

# Example usage:
double_number("a")

```

## If statements (bonus)

An if statement in R is a way to make decisions in your code. It allows you to execute certain code only if a specific condition is true. This is useful when you want your program to behave differently depending on the input or situation.

```{r ifstatement}


target = "sun"

weather = "rainy"


if(target != weather){
  
  print("no sun today :( ")
  
}


```

We can add lots of conditions!

We do this using if else (.... else, else etc.)

```{r ifstatementong}


target = "sun"

weather = "sun"

if(target != weather){
  
  print("no sun today :( ")
  
} else if (target == weather){
  
  print("Sun today :D")
  
}


```

Another example of an if statement in a function with a warning!

```{r ifelsenuma}


check_number <- function(x) {
  if (!is.numeric(x)) {
    warning("Please enter a number.")
  } else if (x > 0) {
    print("The number is positive.")
  } else if (x < 0) {
    print("The number is negative.")
  } else {
    print("The number is zero.")
 }
}


check_number(0)

check_number(1)

check_number(-1)

check_number("hello")



```

Finally we can save the output to a new object

```{r ifelsenum}


check_number <- function(x) {
  if (!is.numeric(x)) {
    warning("Please enter a number.")
  } else if (x > 0) {
    print("The number is positive.")
  } else if (x < 0) {
    print("The number is negative.")
  } else {
    print("The number is zero.")
 }
}

output <- check_number(1)

output


```

## Loops!

A `loop` in R is a programming structure that allows you to repeat a block of code multiple times, which is especially useful when performing repetitive tasks. Instead of writing the same code over and over, you can use a loop to automate the repetition. The most common types of loops in R are `for` loops and `while`loops. A `for` loop runs a set number of times, typically over a sequence like a vector or a range of numbers. A `while` loop continues running as long as a specified condition remains true. Loops are powerful tools for tasks like processing data, generating output, or applying the same operation to multiple items—making your code more efficient, concise, and easier to manage.

```{r simpleloop}


for (i in 1:5) {
  
 print(i)
  
}

```

```{r whilesimple}


count <- 1

while (count <= 8) {
  
  print(count)
  
  count <- count + 1
  
}

```
We can put anything in a loop, for example:

```{r nameloop}

# Vector of names
names <- c("Alice", "Ben", "Charlie", "Diana")

# Loop through each name and create a custom greeting
for (name in names) {
  
  greeting <- paste("Hello", name, "- welcome to the plankton workshop!")
  print(greeting)
  
  # you can put anything in here! make is as long as you lil
  
  
  
  
  
  
  
  
  
  
  
  
  
  # STUFF
  
}


```

## TryExcempt

```{r try}


# Create a test vector with a problematic element
test_vector <- list(10, 5, "error", 2)

# Loop with tryCatch to handle errors
for (val in test_vector) {
  
  result <- tryCatch({
    
    sqrt(val)  # Will fail on the "error" string
    
  }, warning = function(w) {
    
    paste("Warning:", w$message)
    
  }, error = function(e) {
    
    paste("Error:", e$message)
    
  })
  
  print(result)
}



```

# Some examples of all

Using mt cars database!

## Function

```{r fundemo}


# Define the function
high_mpg_cars <- function(threshold) {
  rownames(mtcars)[mtcars$mpg > threshold]
}

# Example usage
high_mpg_cars(20)


```


## If else

```{r ifdemo}


# Select a specific car's mpg value
mpg_value <- mtcars["Mazda RX4", "mpg"]

# Classify the car based on mpg
if (mpg_value > 25) {
  category <- "very fuel efficient"
} else if (mpg_value > 20) {
  category <- "moderately fuel efficient"
} else {
  category <- "not fuel efficient"
}

# Print the result
print(paste("Mazda RX4: ", mpg_value, " mpg - ", category))


```

## Loop

```{r loopdemp}


# Load the mtcars dataset
data(mtcars)

# Loop through each row
for (i in 1:nrow(mtcars)) {
  
  model <- rownames(mtcars)[i]
  mpg <- mtcars$mpg[i]
  
  if (mpg > 20) {
    
    status <- "fuel efficient"
    
  } else {
    
    status <- "not fuel efficient"
    
  }
  
  print(paste(model, ": ", mpg, " mpg - This car is ", status, "."))
}


```

## All three

```{r example}

# Define the function
label_above_mean <- function(data, colname) {
  
  # empty vector
  result <- c()
  
  # for each value in the col
  for (val in data[[colname]]) {
    
    aroundthemean <- if(val > mean(data[[colname]])) {
      "Above Mean"
    } else {
      "Below or Equal"
    }
    
    result <- c(result, aroundthemean)
    
    
  }
  return(result)
}

# Example usage with mtcars and the 'mpg' column
label_above_mean(mtcars, "mpg")

# put as a new column

mtcarsnew <- mtcars

mtcars$aroundmean <- label_above_mean(mtcars, "mpg")

```

# Tasks

🔁 Loops
Sum all even numbers from 1 to 100
Use a for loop to iterate through numbers and add only the even ones to a total.

```{r task1}

```

Print each column name in the mtcars dataset
Loop through colnames(mtcars) and print each name on a new line.

```{r task2}

```

🧠 If Statements
Check if a number is positive, negative, or zero
Write a simple if, else if, else block to classify a number.

```{r task3}

```

Flag cars with mpg over 30 in mtcars
Use an if statement inside a loop to print a message for each car that meets the condition.

```{r task4}

```

🧰 Functions
Create a function that converts Celsius to Fahrenheit
Takes one argument and returns the converted temperature.

```{r task5}

```

Write a function that returns the square of a number
Simple math function that demonstrates input/output.

```{r task6}

```
