---
title: "Basic Statistics in R"
author: "Teresa Vaz Martins and Karolina Zalewska"
output: 
  html_document:
    theme: flatly 
    code_folding: show
    #toc: true
    #toc_float: true
editor_options: 
  markdown: 
    wrap: 80
    canonical: true
date: ""
---

```{css, echo=FALSE}

.lightgrey {
background-color:#F6F6F6;
color: #000; 
font-size:16px; 
text-align:left;
vertical-align: middle;
border-radius: 5px;
padding: 20px; 
80px

}

body{
font-family: Helvetica;
font-size: 14pt;
}


```

# Introduction

In this session, we will give you a brief introduction and an overview of how to
perform some basic statistics in R.

**R is a statistical programming language:** by default, it includes many
statistical functions. These are held in the {stats} package, which is
pre-installed with R.

# Exploratory data analysis - summary statistics

Before you run any tests, you should familiarise yourself with the data! Some
things to think about: - how many variables are there? what type of data are
they? check the *class* of each variable and make sure that all data is coded as
expected (e.g. categorical variables are factors) - calculate some summary
statistics of the variables (mean, minimum, maximum), plot histograms. Does
everything look as expected? - how many observations are there? are there any
missing values? how are missing values coded in your data?

We are going to have a quick look at some data using functions you should
already be mostly familiar with.

First we will load a dataset that is stored within a package. The data were
collected and made available by Dr. Kristen Gorman and the Palmer Station,
Antarctica LTER [palmerpenguins
(allisonhorst.github.io)](https://allisonhorst.github.io/palmerpenguins/index.html)

Below are a few examples of things you might do to explore your data. This is by
no means an exhaustive list - the methods you use to explore your data will be
dependent on the data itself (e.g. the type of variables present). While I look
at only two variables, you should always check all variables that you are
planning to include in your analysis.

```{r EDA, eval = FALSE}

# install.packages("palmerpenguins") # uncomment (remove '#' at the start of the line) only if {palmerpenguins} not installed on your machine
library(palmerpenguins)

str(penguins) # the structure function gives a summary of the dataset

# check for missing values
# is.na(penguins$bill_length_mm) # the output of this will be a boolean (true/false) vector of the same size as there are rows in the dataset
table(is.na(penguins$bill_length_mm))

# summary statistics 
mean(penguins$bill_length_mm) # what is the issue here?
mean(penguins$bill_length_mm, na.rm = TRUE) # remove NAs from mean calculation

summary(penguins$bill_length_mm)

hist(penguins$bill_length_mm)


class(penguins$species)

# species is a categorical variable, we can see how many species are in the dataset (factor levels) and how many observations per species
unique(penguins$species) # list unique values of variable
length(unique(penguins$species)) #
table(penguins$species) 

```

There are also more efficient ways of checking your data - have a look at the
materials from other sessions of this course. Additionally, as a starting point
and to get ideas on what to explore, you can use packages such as
{DataExplorer}. This will give you a summarised report of your data.

```{r DataExplorer, eval = FALSE}

# install.packages("DataExplorer")

DataExplorer::create_report(penguins)

```

**Further reading:**

[Simple Fast Exploratory Data Analysis in R with DataExplorer Package\|by
AbdulMajedRaja RS \| Towards Data
Science](%5Bhttps://towardsdatascience.com/simple-fast-exploratory-data-analysis-in-r-with-dataexplorer-package-e055348d9619)](<https://towardsdatascience.com/simple-fast-exploratory-data-analysis-in-r-with-dataexplorer-package-e055348d9619>))

# Parametric and non-parametrics tests

**What is "parametric"?**

statistical tests come with assumptions that need to be met for any inference
drawn from them to be valid. In particular, parametric tests make an assumption
about the distribution of the data

**A quick note on hypothesis testing** one-way vs two-way

## Comparing groups

### I want to compare my data to a single value: 1-sample t-test / Wilcoxon rank test

### I want to compare before and after data: paired t-test / Wilcoxon rank test

### I want to compare values between two independent groups: (unpaired) t-test / Mann Whitney U test

### I want to compare more than two independent groups: ANOVA / Kruskal-Wallis

## Association: chi-square

## Relationship: correlation

# Linear model

## single explanatory variable

## more than one explanatory variable

## assumptions
