---
title: "Basic Statistics in R"
author: "Teresa Vaz Martins and Karolina Zalewska"
output: 
  html_document:
    theme: flatly 
    code_folding: show
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 80
    canonical: true
date: ""
---

```{r setup, include=FALSE}
library(knitr)
# library(see)
library(DataExplorer)
library(palmerpenguins)
library(ggplot2)
library(ggeffects)
library(car)
library(performance)
library(corrplot)

## if any of these produce an error message use install.packages(), e.g. 
# install.packages(c("knitr", "DataExplorer", "palmerpenguins", "ggplot2",
#                    "ggeffects", "car", "performance", "corrplot"))
# (delete as needed)

```

```{css, echo=FALSE}

.lightgrey {
background-color:#F6F6F6;
color: #000; 
font-size:14px; 
text-align:left;
vertical-align: middle;
border-radius: 5px;
padding: 20px; 
80px

}

body{
font-family: Helvetica;
font-size: 12pt;
}
 

```

# Introduction

In this session, we will give you a brief introduction and an overview of how to
perform some basic statistics in R.

**R is a statistical programming language:** by default, it includes many
statistical functions. These are held in the {stats} package, which comes
pre-installed with R.

# Exploratory data analysis - summary statistics

Before you run any tests, you should familiarise yourself with the data! Some
things to think about:

<ul>

<li>how many variables are there? what type of data are they? check the *class*
of each variable and make sure that all data is coded as expected (e.g.
categorical variables are factors)</li>

<li>calculate some summary statistics of the variables (mean, minimum, maximum),
plot histograms. Does everything look as expected?</li>

<li>how many observations are there? are there any missing values? how are
missing values coded in your data? How should they be treated/could they be
removed?</li>

</ul>

We are going to have a quick look at some data using functions you should
already be mostly familiar with.

First we will load a dataset that is stored within a package. The data were
collected and made available by Dr. Kristen Gorman and the Palmer Station,
Antarctica LTER [palmerpenguins
(allisonhorst.github.io)](https://allisonhorst.github.io/palmerpenguins/index.html)

Below are a few examples of things you might do to explore your data. This is by
no means an exhaustive list - the methods you use to explore your data will be
dependent on the data itself (e.g. the type of variables present). While I look
at only two variables, you should always check all variables that you are
planning to include in your analysis.

```{r EDA}

str(penguins) # the structure function gives a summary of the dataset

# check for missing values
# is.na(penguins$bill_length_mm) # the output of this will be a boolean 
# (true/false) vector of the same size as there are rows in the dataset
table(is.na(penguins$bill_length_mm))

# summary statistics 
mean(penguins$bill_length_mm) # what is the issue here?
mean(penguins$bill_length_mm, na.rm = TRUE) # remove NAs from mean calculation

summary(penguins$bill_length_mm)

hist(penguins$bill_length_mm)


class(penguins$species)

# species is a categorical variable, we can see how many species are in the 
# dataset (factor levels) and how many  observations per species
unique(penguins$species) # list unique values of variable
length(unique(penguins$species)) #
table(penguins$species) 

```

Additionally, as a starting point and to get ideas on what to explore, you can
use packages such as {DataExplorer}. This will give you a summarised report of
your data.

```{r DataExplorer, eval = FALSE}

DataExplorer::create_report(penguins)

```

**Further reading:**
<a href = "https://towardsdatascience.com/four-r-packages-for-automated-exploratory-data-analysis-you-might-have-missed-c38b03d4ee16/">Four
R packages for Automated Exploratory Data Analysis you might have missed</a>

# Parametric and non-parametrics tests

## Comparing groups

### I want to compare my data to a single value: 1-sample t-test / Wilcoxon rank test

We want to compare populations of Adelie penguins on different islands, and test
whether there is a difference in bill length across the populations.

What if we had individual measurements for one island and wanted to know how it
differs from an average Adelie penguin bill length that we found from
literature?

```{r 1-sample test}

# extract the data from the penguins dataset
a1 <- penguins[penguins$species == "Adelie" & penguins$island == "Dream", ] 

mu <- 38.79 # this is our average bill length obtained from literature

t.test(a1$bill_length_mm, alternative = "two.sided", mu = mu) 

# if our H was that the population's bill length is shorter than average
# t.test(a1$bill_length_mm, alternative = c("less"), mu = mu) 
# if our H was that the population's bill length is longer than average
# t.test(a1$bill_length_mm, alternative = c("greater"), mu = mu) 

# is our data normally distributed?
hist(a1$bill_length_mm)

# we can perform a Shapiro-Wilk test to find out if data follows normal distribution
# H0: distribution of sample data no different to normal distribution

# NOTE: the Shapiro-Wilk test is a TEST so the same caveats and potential pitfalls 
# of any statistical testing also apply to it. It can be useful but always
# interpret with caution and always in addition to other ways of checking
# normality. Also, it is not useful for large samples: the maximum sample size
# is only 5,000 observations. 
shapiro.test(a1$bill_length_mm)


### NON-PARAMETRIC ALTERNATIVE

wilcox.test(a1$bill_length_mm, alternative = "two.sided", mu = mu)


```

### I want to compare two independent groups: (unpaired) t-test / Mann Whitney U test

This time, we went to two different islands and collected individual
measurements of Adelie penguin bill lengths. We want to test whether there is a
difference between the two populations inhabiting the different islands.

```{r two-sample t-test, warning = FALSE, warning = FALSE, messgae = FALSE}

# extract the data from the penguins dataset
a1 <- penguins[penguins$species == "Adelie" & penguins$island == "Dream", ]
a2 <- penguins[penguins$species == "Adelie" & penguins$island == "Torgersen", ]


t.test(x = a1$bill_length_mm, y = a2$bill_length_mm, 
       alternative = "two.sided", 
       var.equal = TRUE) 
# the var.equal argument -> are we assuming equal variance between the two samples?


## check if variances equal: F-test
var.test(x = a1$bill_length_mm, y = a2$bill_length_mm, 
         alternative = "two.sided")

# unequal variance between the samples? -> welch's t-test 
# t.test(a1$bill_length_mm, a2$bill_length_mm, alternative = "two.sided", 
# var.equal = FALSE)

# check normality assumption
a <- rbind(a1, a2)

ggplot(data = a) +
  geom_histogram(aes(x = bill_length_mm), fill = "white", color = "black") +
  facet_wrap(~island) +
  xlab("Bill length [mm]") +
  theme_classic()


## NON-PARAMETRIC
wilcox.test(x = a1$bill_length_mm, y = a2$bill_length_mm,
            alternative = "two.sided")

```

### I want to compare before and after data: paired t-test / Wilcoxon rank test

In this example we are responsible for the management of multiple sites, on
which we carry out regular surveys of a species of conservation concern. We want
to know if a recently implemented change in land management had an effect on
individual counts. To do that, we repeat the survey after the management has
changed and compare our counts to those collected before the change.

```{r paired t-test}

set.seed(123) 
# I am generating random numbers, running the above means that we get the same
# numbers every time

## create the example dataset
df <- data.frame(site = 1:100,
                 before = round(runif(n = 100, min = 0, max = 10)), 
                 after = round(runif(n = 100, min = 0 , max = 10)))


head(df)

# NOTE: 
# runif() generates n random samples from the uniform distribution. 
# You can get samples from e.g. the Normal distribution using rnorm(), 
# and equivalent functions exist for other distributions. 
# round() rounds the values to a specified number of decimal places, by default 0
# ceiling() and floor() round to integer (either "up" or "down")

t.test(x = df$before, y = df$after, alternative = "two.sided",
       paired = TRUE, var.equal = TRUE)


# NON-PARAMETRIC ALTERNATIVE
wilcox.test(x = df$before, y = df$after, alternative = "two.sided",
            paired = TRUE)


```

### I want to compare more than two independent groups: ANOVA / Kruskal-Wallis

We've not seen any differences in bill lengths within the Adelie penguins on
different islands, are there bill lengths different between the three species?

```{r anova}

anova_model <- aov(formula = bill_length_mm ~ species, data = penguins)

summary(anova_model)

# you could plot the model to check if it meets some ANOVA assumptions 
# (eg normality) - we will be discussing those plots later
# plot(anova_model)

# NON-PARAMETRIC ALTERNATIVE
kruskal.test(x = penguins$bill_length_mm, g = penguins$species)
kruskal.test(formula = bill_length_mm ~ species, data = penguins)


```

The low p-value ($<2 \times 10^{-16}$) indicates strong evidence against the
null hypothesis, suggesting that at least one group mean is significantly
different from the others. The F-statistic measures the ratio of the variance
between groups to the variance within groups. A higher F-value indicates a
greater difference between group means.

## I want to test the association between factors or compare counts: chi-square test

Chi-square (or $\chi^2$) test is used to check for association, or compare
counts across multiple groups. The test is formulated via a contingency table.
Imagine you carried out a survey of three species (1-3) across three field sites
(A-C). The contingency table may look something like this:

```{r chi-square table, results = 'asis'}

df <- data.frame(sp1 = c(16, 8, 22, 46),
                 sp2 = c(5, 30, 12, 47),
                 sp3 = c(21, 19, 6, 46),
                 row.names = c("A", "B", "C", "column_total"))


df$row_total <- rowSums(df)

kable(df, format = "html", align = "c")


```

<br>

You want to know whether a species occurs at any site more frequently.

```{r chisquare}

df <- data.frame(sp1 = c(16, 8, 22),
                 sp2 = c(5, 30, 12),
                 sp3 = c(21, 19, 6),
                 row.names = c("A", "B", "C"))

res <- chisq.test(df)
res

# the test statistic is calculated using the observed and expected values, have a look at these
res$observed
res$expected

# we can use the residuals to further interpret the results
res$residuals

```

Residuals sign: positive = observed \> expected; negative = observed \< expected

absolute value = magnitude of difference between observed and expected

## Relationship: correlation

Are penguin bill lengths and depths correlated?

```{r correlation coefficient}
## linear and normally distributed - pearson

cor(x = penguins$bill_length_mm, y = penguins$bill_depth_mm, use = "complete.obs",
    method = "pearson")

cor.test(x = penguins$bill_length_mm, y = penguins$bill_depth_mm,
         alternative = "two.sided", method = "pearson")


## non-linear and not normally distributed = spearman
cor(x = penguins$bill_length_mm, y = penguins$bill_depth_mm, use = "complete.obs",
    method = "spearman")

```

Interpretation:

<ul>

<li>sign = positive/negative correlation</li>

<li>value: the values of the correlation coefficient are [-1, 1], the closer to
\|1\| the stronger the correlation</li>

</ul>

# Linear regression

Linear regression models the relationship between a dependent variable and one
or more independent variables. As a reminder, below is the equation of a
straight line: <br>

$y = mx + c$

Where:

<ul>

<li>m is the slope (change in y per 1 unit increase in x)</li>

<li>c is the y-interecept (i.e. where x = 0)</li>

</ul>

<br>

**Quick note on terminology:** I use the below terms interchangeably:

<ul>

<li>response variable = outcome variable = dependent variable = y</li>

<li>explanatory variable = predictor variable = independent variable = x</li>

</ul>

## single explanatory variable

$y = \beta_0 + \beta_1 x_1 + \epsilon$

```{r LM1}

mod1 <- lm(bill_length_mm ~ species, data = penguins)

summary(mod1)

```

The sign of the coefficient (positive or negative) indicates the direction of
the relationship, while the magnitude reflects the strength of the relationship.
R-squared (coefficient of determination) gives the proportion of variance within
the response variable that is explained by the explanatory variable. Including
more explanatory variables will increase the R-squared, however, this may not be
because of true relationships between those variables and the response, but just
chance. An attempt to prevent that is the adjusted R-squared, which is
calculated taking into account the number of explanatory variables (predictors).

## more than one explanatory variable (multiple linear regression)

$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \epsilon$

```{r LM2}

mod2 <- lm(bill_length_mm ~ species + flipper_length_mm + sex, data = penguins)

summary(mod2)

```

## predicting from model

We can quickly plot the modelled relationship by predicting from the model. We
obtain a predicted bill length for a given species, flipper length, and sex,
from our model using the formula: <br>
$bill \ length = \beta_0 + \beta_1 \times species + \beta_2 \times flipper \ length + \beta_3 \times sex$.
<br>

```{r predicting}
pred <- ggpredict(mod2, terms = "species", ci_level = 0.95)

ggplot(data = pred, aes(x = x, y = predicted)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1) +
  labs(x = "Species", y = "Predicted bill length [mm]") +
  theme_classic()


pred <- ggpredict(mod2, terms = "flipper_length_mm", ci_level = 0.95)

ggplot(data = pred, aes(x = x, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "grey50", alpha = 0.4) +
  labs(x = "Filpper length [mm]", y = "Predicted bill length [mm]") +
  theme_classic()

```

```{r extra predicting, eval = FALSE, include = TRUE}
## ADDITIONAL INFO
## By default, ggpredict() will predict across the range of the variable, 
## you can also get prediction for specific value(s). 

## e.g. if you wanted to predict for two species of penguin only:
ggpredict(mod2, terms = "species [Gentoo, Adelie]", ci_level = 0.95)

## or for a specifc range of values
p1 <- ggpredict(mod2, terms = "flipper_length_mm [200:300]", ci_level = 0.95)

ggplot(data = p1, aes(x = x, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "grey50", alpha = 0.4) +
  labs(x = "Flipper length [mm]", y = "Predicted bill length [mm]") +
  theme_classic()

## if you want to make sure that the function is predicting for all values of the
## variable, you can use:
# ggpredict(mod2, terms = "species [all]", ci_level = 0.95)

```

## Interaction

$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2 + \epsilon$

**Note formula** in R you can specify an interaction in two equivalent ways (the
output will be exactly the same):

$y \sim x_1 * x_2$ OR $y \sim x_1 + x_2 + x_1:x_2$

Unless you know what you're doing and have a good reason, your model needs to
contain both the lower order (additive, i.e. $x_1 + x_2$) and higher order
(interaction) terms.

```{r LM3}

mod3 <- lm(bill_length_mm ~ flipper_length_mm * species + sex, data = penguins)

summary(mod3)


# I prefer to use '*' to specify interactions, but I've left this here if you 
# would like to check that both methods give the same output

# mod3 <- lm(bill_length_mm ~ species + flipper_length_mm + flipper_length_mm:species,
#            data = penguins)

# summary(mod3)


```

Interpretation of the interaction terms gets slightly trickier, but the general
idea is the same. However, the easiest way to figure out what is happening is to
plot it. We will again use our model (mod3) to get the predicted bill length for
a range of flipper length values across the three species.

```{r pred mod with int}
pred <- ggpredict(mod3, terms = c("flipper_length_mm", "species"), ci_level = 0.95)

ggplot(data = pred, aes(x = x, y = predicted, fill = group)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(x = "Flipper length [mm]", y = "Predicted bill length [mm]") +
  scale_fill_brewer(name = "Species", palette = "Dark2") +
  theme_classic()


```

## Checking model assumptions

Below are the assumptions of a linear model:

<ol>

<li>Linearity: the relationship between the response variable and the
explanatory variables is linear</li>

<li>Homoscedasticity: the variance is equal across range/all levels of
explanatory (independent) variable(s)</li>

<li>Normality:the errors are normally distributed</li>

<li>Independence of observations: no residual correlation</li>

</ol>

```{r assumptions, fig.width=12, fig.height=12}

par(mfrow = c(2, 2))
plot(mod2)


check_model(mod2)


# multicollinearity

## variance inflation factor
vif(mod2)

## you can also make a correlation matrix and plot to check how correlated 
## continuous variables are
## N.B. in the models above we only ever include one of these variables so the 
## below is not relevant for our example models
mtr <- cor(penguins[ , c("body_mass_g", "bill_depth_mm", "flipper_length_mm")],
           use = "complete.obs")
mtr

corrplot(mtr, method = "ellipse", type = "lower")
```

## Try it yourself: linear models

You want to find out what affects the amount of sleep of mammals. To do that,
you will use data for 83 species ('msleep'). Load the data, inspect the dataset,
consider which variables should be included in the model, and fit a linear
model. Check model assumptions and plot your results

```{r exercise}

data("msleep") # if you can't find the data, make sure you load the 'ggplot2' package



```
