---
title: "Basic Statistics in R"
author: "Teresa Vaz Martins and Karolina Zalewska"
output: 
  html_document:
    theme: flatly 
    code_folding: show
    #toc: true
    #toc_float: true
editor_options: 
  markdown: 
    wrap: 80
    canonical: true
date: ""
---

```{css, echo=FALSE}

.lightgrey {
background-color:#F6F6F6;
color: #000; 
font-size:16px; 
text-align:left;
vertical-align: middle;
border-radius: 5px;
padding: 20px; 
80px

}

body{
font-family: Helvetica;
font-size: 14pt;
}


```

# Introduction

In this session, we will give you a brief introduction and an overview of how to
perform some basic statistics in R.

**R is a statistical programming language:** by default, it includes many
statistical functions. These are held in the {stats} package, which is
pre-installed with R.

# Exploratory data analysis - summary statistics

Before you run any tests, you should familiarise yourself with the data! Some
things to think about: - how many variables are there? what type of data are
they? check the *class* of each variable and make sure that all data is coded as
expected (e.g. categorical variables are factors) - calculate some summary
statistics of the variables (mean, minimum, maximum), plot histograms. Does
everything look as expected? - how many observations are there? are there any
missing values? how are missing values coded in your data?

We are going to have a quick look at some data using functions you should
already be mostly familiar with.

First we will load a dataset that is stored within a package. The data were
collected and made available by Dr. Kristen Gorman and the Palmer Station,
Antarctica LTER [palmerpenguins
(allisonhorst.github.io)](https://allisonhorst.github.io/palmerpenguins/index.html)

Below are a few examples of things you might do to explore your data. This is by
no means an exhaustive list - the methods you use to explore your data will be
dependent on the data itself (e.g. the type of variables present). While I look
at only two variables, you should always check all variables that you are
planning to include in your analysis.

```{r EDA, eval = FALSE}

# install.packages("palmerpenguins") # uncomment (remove '#' at the start of the line) only if {palmerpenguins} not installed on your machine
library(palmerpenguins)

str(penguins) # the structure function gives a summary of the dataset

# check for missing values
# is.na(penguins$bill_length_mm) # the output of this will be a boolean (true/false) vector of the same size as there are rows in the dataset
table(is.na(penguins$bill_length_mm))

# summary statistics 
mean(penguins$bill_length_mm) # what is the issue here?
mean(penguins$bill_length_mm, na.rm = TRUE) # remove NAs from mean calculation

summary(penguins$bill_length_mm)

hist(penguins$bill_length_mm)


class(penguins$species)

# species is a categorical variable, we can see how many species are in the dataset (factor levels) and how many observations per species
unique(penguins$species) # list unique values of variable
length(unique(penguins$species)) #
table(penguins$species) 

```

There are also more efficient ways of checking your data - have a look at the
materials from other sessions of this course. Additionally, as a starting point
and to get ideas on what to explore, you can use packages such as
{DataExplorer}. This will give you a summarised report of your data.

```{r DataExplorer, eval = FALSE}

# install.packages("DataExplorer")

DataExplorer::create_report(penguins)

```

**Further reading:**

[Simple Fast Exploratory Data Analysis in R with DataExplorer Package\|by
AbdulMajedRaja RS \| Towards Data
Science](%5Bhttps://towardsdatascience.com/simple-fast-exploratory-data-analysis-in-r-with-dataexplorer-package-e055348d9619)](<https://towardsdatascience.com/simple-fast-exploratory-data-analysis-in-r-with-dataexplorer-package-e055348d9619>))

# Parametric and non-parametrics tests

**What is "parametric"?**

statistical tests come with assumptions that need to be met for any inference
drawn from them to be valid. In particular, parametric tests make an assumption
about the distribution of the data

**A quick note on hypothesis testing** one-way vs two-way

## Comparing groups

### I want to compare my data to a single value: 1-sample t-test / Wilcoxon rank test

### I want to compare before and after data: paired t-test / Wilcoxon rank test

### I want to compare values between two independent groups: (unpaired) t-test / Mann Whitney U test

### I want to compare more than two independent groups: ANOVA / Kruskal-Wallis

The variable \`Species\` can be seen as an independent variable to conduct an
ANOVA test. This is a statistical test to check three or more differences in
group means. The code snippet below performs an ANOVA test using the
`palmerpenguins` dataset (specified by the `data` parameter), specifically, we
will compare the bill length (measured in mm) among the species. (`formula`
parameter defines the relationship to be tested).

```{r}

anova_model <- aov(formula = bill_length_mm ~ species, data = penguins)

summary(anova_model)

#' you could plot the model to check if it meets some ANOVA assumptions (eg normality)
# plot(anova_model)
```

The low p-value ($<2 \times 10^{-16}$) indicates strong evidence against the
null hypothesis, suggesting that at least one group mean is significantly
different from the others. The F-statistic measures the ratio of the variance
between group means to the variance within groups. A higher F-value indicates a
greater difference between group means.

## Association: chi-square

## Relationship: correlation

# Linear regression

$y = mx + c$

## single explanatory variable

$y = \beta_0 + \beta_1 x_1 + \epsilon$

```{r LM1}

mod1 <- lm(bill_length_mm ~ species, data = penguins)

summary(mod1)

```

## more than one explanatory variable (multiple linear regression)

$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + ... + \epsilon$

```{r LM2}

mod2 <- lm(bill_length_mm ~ species + body_mass_g + sex, data = penguins)

summary(mod2)

```

## predicting from model

We can quickly plot the modelled relationship by predicting from the model. We
could obtain a predicted bill length for a given species, body mass, and year,
from our model using the formula
$bill \ length = \beta_0 + \beta_1 \times species + \beta_2 \times body \ mass + \beta_3 \times sex$.
If we want plot the relationship between

```{r predicting}
library(ggplot2)
library(ggeffects)

pred <- ggpredict(mod2, terms = "species", ci_level = 0.95)

ggplot(data = pred, aes(x = x, y = predicted)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1) +
  labs(x = "Species", y = "Predicted bill length [mm]") +
  theme_classic()


pred <- ggpredict(mod2, terms = "body_mass_g", ci_level = 0.95)

ggplot(data = pred, aes(x = x, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "grey50", alpha = 0.4) +
  labs(x = "Body mass [g]", y = "Predicted bill length [mm]") +
  theme_classic()

```

```{r extra predicting, eval = FALSE}
## ADDITIONAL INFO
## By default, ggpredict() will predict across the range of the variable, 
## you can also get prediction for specific value(s). 

## e.g. if you wanted to predict for two species of penguin only:
ggpredict(mod2, terms = "species [Gentoo, Adelie]", ci_level = 0.95)

## or for a specifc range of values
p1 <- ggpredict(mod2, terms = "body_mass_g [3000:4000]", ci_level = 0.95)

ggplot(data = p1, aes(x = x, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "grey50", alpha = 0.4) +
  labs(x = "Body mass [g]", y = "Predicted bill length [mm]") +
  theme_classic()

## if you want to make sure that the function is predicting for all values of the
## variable, you can use:
# ggpredict(mod2, terms = "species [all]", ci_level = 0.95)

```

## Interaction

$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2 + \epsilon$

**Note formula** in R you can specify an interaction in two equivalent ways (the
output will be exactly the same): $y \sim x_1 * x_2$ OR
$y \sim x_1 + x_2 + x_1:x_2$ Unless you know what you're doing and have a good
reason, your model needs to contain both the lower order (additive, i.e.
$x_1 + x_2$) and higher order (interaction) terms.

```{r LM3}

mod3 <- lm(bill_length_mm ~ flipper_length_mm * species + sex, data = penguins)

summary(mod3)


# I prefer to use '*' to specify interactions, but I've left this here if you 
# would like to check that both methods give the same output

# mod3 <- lm(bill_length_mm ~ species + flipper_length_mm + flipper_length_mm:species,
#            data = penguins)

# summary(mod3)


```

Interpretation of the interaction terms gets slightly trickier, but the general
idea is the same. However, the easiest way to figure out what is happening is to
plot it. We will again use our model (mod3) to get the predicted bill length for
a range of flipper length values across the three species.

**Important** do not interpret the lower order terms (main effects) when higher
order terms present, i.e. in our case, because we have an interactive effect of
body mass and species we should not be interpreting the coefficients related to
species or body mass.

```{r pred mod with int}
# library(ggplot2)
# library(ggeffects)

pred <- ggpredict(mod3, terms = c("flipper_length_mm", "species"), ci_level = 0.95)

ggplot(data = pred, aes(x = x, y = predicted, fill = group)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  labs(x = "Flipper length [mm]", y = "Predicted bill length [mm]") +
  scale_fill_brewer(name = "Species", palette = "Dark2") +
  theme_classic()


```

## Checking model assumptions

```{r assumptions}

# install.packages(car)
# install.packages(performance)
# install.packages(corrplot)

library(car)
library(performance)
library(corrplot)


par(mfrow = c(2, 2))
plot(mod2)


check_model(mod2)



# multicollinearity

## variance inflation factor
vif(mod2)

## you can also make a correlation matrix and plot to check how correlated 
## continuous variables are
## N.B. in the models above we only ever include one of these variables so the 
## below is not relevant for our example models
mtr <- cor(penguins[ , c("body_mass_g", "bill_depth_mm", "flipper_length_mm")],
           use = "complete.obs")
mtr

corrplot(mtr, method = "ellipse", type = "lower")
```
