---
title: "Advanced statistics in R"
author: "Teresa Vaz Martins and Karolina Zalewska"
output: 
  html_document:
    theme: flatly 
    code_folding: show
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 80
    canonical: true
date: ""
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(factoextra)
library(lme4)
library(MASS)
library(ggeffects)
library(mgcv)
library(performance)
library(palmerpenguins)
library(gamair)
library(ggpubr)

```

```{css, echo=FALSE}

.lightgrey {
background-color:#F6F6F6;
color: #000; 
font-size:16px; 
text-align:left;
vertical-align: middle;
border-radius: 5px;
padding: 20px; 
80px

}

body{
font-family: Helvetica;
font-size: 14pt;
}


```

# Dimensionality reduction

## Principal component analysis (PCA)

To visualize high-dimensional data, we use PCA to map data to lower dimensions.
The method is particularly useful when the variables within the data set are
highly correlated. Due to this redundancy, PCA can be used to reduce the
original variables into a smaller number of new variables ( = *principal
components*). These will not be a reduced portion of the original variables, it
will be different variables!

We can do PCA in R in different ways, for instance

\- `prcomp`() [built-in R `stats` package],

\- `PCA`() [`FactoMineR` package, useful for visualisation],

**Note:** run `?prcomp` on the console to check for other possible arguments,
besides `x`

Of the possible arguments, note the `scale` and `center` arguemnts. These
standardise the input data so that it has a mean of zero and variance of one.
This is particularly recommended when variables are measured in different scale
(e.g. different units: kilograms, grams). By

```{r PCA1, error=TRUE}

# make sure to use numeric data and remove missing values before running PCA
df <- iris[ , c("Sepal.Length", "Sepal.Width",
                             "Petal.Length", "Petal.Width")]


df <- na.omit(df)

iris_pca <- prcomp(df, scale = TRUE, center = TRUE)


iris_pca

```

**Note:** The `rank.` argument allows us to select the number of principal
components to retain.

### Visualise results

Although the default `stats` package includes options to produce many PCA
visualisations, I find it easier to customise `factoextra`. We will run just a
few examples, but you can find other options online, for instance here:

[PCA - Principal Component Analysis Essentials - Articles
-STHDA](%5Bhttp://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/)](<http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/>))

#### Scree plot and variable contributions

Visualizing the variance explained by each component helps us to visually
identify how many principal components are needed.

```{r scree, warning = FALSE}

#calculate total variance explained by each principal component
var_explained <- iris_pca$sdev^2 / sum(iris_pca$sdev^2)

fviz_eig(iris_pca, addlabels = TRUE, ylim = c(0, 50))


# Contributions of variables to PC1

fviz_contrib(iris_pca, choice = "var", axes = 1)

# Contributions of variables to PC2
fviz_contrib(iris_pca, choice = "var", axes = 2)


fviz_pca_var(iris_pca, axes = c(1, 2), col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE) +
  xlab("PC1") +
  ylab("PC2") +
  theme(text = element_text(size = 14))



```

```{r}

fviz_pca_ind(iris_pca ,
             geom.ind = "point", # show points
             col.ind = iris$Species, # color by species, iris$Species
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Groups")


```

**Further reading:**

<ul>

<li>PCA using the [recipes](https://recipes.tidymodels.org) package from
[tidymodels](https://www.tidymodels.org)</li>

<li>[PCA with penguins and
recipes](https://allisonhorst.github.io/palmerpenguins/articles/pca.html)</li>

<li>Using the factoextra package: [PCA - Principal Component Analysis
Essentials -
Articles-STHDA](http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials)</li>

<li>[PCA with tidyverse - Benjamin
Nowak](https://bjnnowak.netlify.app/2021/09/15/r-pca-with-tidyverse)</li>

</ul>

# Going beyond the Gaussian error distribution: Generalized linear models (GLM)

These can be used to model data that is not normally distributed. The different
types are based on the probability distribution of the response variable: e.g.
Poisson if the response variable represents counts or binomial for a binary
outcome.

GLMs are fit using the `glm()` function. The formula is specified in the same
way as for `lm()`, but there is an additional argument to pay attention to: <br>

**family:** The statistical family to use to fit the model. Default is gaussian,
as in a linear model, but other options include binomial, Gamma, and poisson
among others. You can find more information on the types of model families
within {stats} using `?family`.

## I have count data / I'm interested in the rate : Poisson distribution

Going back to our site management example: we are responsible for the management
of multiple sites, which are regularly surveyed. On each survey, the number of
individuals observed is recorded. The sites a mixture of forested and open
habitats, and for each site we have the average percentage of tree cover. We
would like to find out if the species is found more frequently on sites with a
greater tree cover.

```{r}

set.seed(123)

df <- data.frame(site = 1:100,
                 count = round(runif(n = 100, min = 0, max = 10)), 
                 trees = runif(n = 100, min = 0 , max = 100))


mod1 <- glm(count ~ trees, family = poisson(link = "log"), data = df)


summary(mod1)

check_model(mod1)

## overdispersion?

mod1q <- glm(count ~ trees, family = quasipoisson(link = "log"), data = df)

summary(mod1q)

mod1nb <- glm.nb(count ~ trees, data = df)

summary(mod1nb)


```

## I have binomial (0/1 outcome) data:

What if we didn't have counts but only a presence/absence of the species?

```{r}

set.seed(123)

df <- data.frame(site = 1:100,
                 count = round(runif(n = 100, min = 0, max = 10)), 
                 trees = runif(n = 100, min = 0 , max = 100))


df <- df %>% 
  mutate(pres = ifelse(count > 4, 1, 0))



mod2 <- glm(pres ~ trees, family = binomial(link = "logit"), data = df)

summary(mod2)

```

The odds ratio for each coefficient represents the average increase in the odds
of an individual defaulting, assuming all other predictor variables are held
constant.

```{r, error=TRUE, message=FALSE}

#calculate odds ratio and 95% confidence interval for each predictor variable 

exp(cbind(Odds_Ratio = coef(mod2), confint(mod2)))
```

**Be careful with interpretation!!**Odds ratio is notorious for not being very
intuitive to interpret. Below, I've listed some potentially useful resources

<ul>

<li>[Introduction to Regression Methods
Ch.6](https://www.bookdown.org/rwnahhas/RMPH/blr-interp.html)</li>

<li>[RPubs Interpreting the Output of a Logistic Regression
Model](https://rpubs.com/raoulbia/interpreting_glm_logistic_regression_output)</li>

</ul>

# Mixed Effects Models in R with lme4

Mixed effects models, also known as multilevel models or hierarchical models,
are statistical models that are used to analyze data with nested or hierarchical
structures. These models are particularly useful when dealing with data that has
multiple levels of variation or when there is a need to account for the
correlation between observations within the same group.

In R, the {lme4} package provides functions for fitting mixed effects models.
The `lmer()` function is commonly used to fit linear mixed effects models

The formula synatx is the same as for all models we have seen so far, however,
we can specify a random effect term. <br>
$y \sim x_1 + x_2 ... + (1|group), data = data$ <br>

where `group` is the random effects variable.

The `(1 | group)` term specifies that the random intercepts that vary across
different levels of the \`group\` variable.

We could also specify a random slope and random intercept term as follows:
$y \sim x_1 + x_2 ... + (1+group \ var_1|group \ var_2), data = data$ <br>

The random effects can be nested: <br>
$y \sim x_1 + x_2 ... + (1|group \ var_1 :group \ var_2), data = data$ <br> or
crossed: <br>
$y \sim x_1 + x_2 ... + (1|group \ var_1) + (1|group \ var_2), data = data$ <br>

The random effect structure to use will dependent on the hierarchical structure
of the data.

**Note**Forums such as StackOverflow or StackExchange are not only good places
to look for code advice but also stats questions and application in R. As an
example, here is a link to a very good answer about the difference between
[crossed and nested random
effects](https://stats.stackexchange.com/questions/228800/crossed-vs-nested-random-effects-how-do-they-differ-and-how-are-they-specified)

Going back to our penguin example: we are still interested in the bill length,
however, we know that the individuals measured come from different islands.
Could there be some differences across these islands that affect the
individuals? Often we have good reason to believe that there may be additional
sources of variation associated, for example, with differences in study sites.
However, it is impossible to reasonably include all sources of variation in the
model. Instead, we can try to account for these by identifying a variable that
"encompasses" them in a way. For example, it may be safe to assume that the
islands are different, what exactly is different about them, we do not know, but
our goal is not to understand how they are different but just to account for
them being potentially different.

Below is the model using species, body mass, and sex to explain bill length of
penguins. We also specify a random intercept effect of island. This means that

```{r, error=TRUE, warning=FALSE, message=FALSE}

#library(lme4) # already loaded at start

mod3 <- lme4::lmer(bill_length_mm ~ species + body_mass_g + sex + (1|island), 
                   data = penguins)

summary(mod3)

```

The random effects terms `(1|island)` allows individuals to vary randomly in
terms of their intercept depending on the island they were found on.

Note that the output above does not give us p-values by default `lmer` from
{lme4} does not calculate these, however, if you do need p-values, you can use
either {lmerTest} or {nlme} to re-fit the model and obtain p-values. The
parameter estimates should be the same, but some packages are built to work with
{lme4} output rather than, e.g.{lmerTest}, so you may need to use both packages
to produce e.g. prediction plots or check model assumptions

```{r, message = FALSE, warning = FALSE}

# install.packages("lmerTest")
library(lmerTest)

mod3p <- lmerTest::lmer(bill_length_mm ~ species + body_mass_g + sex + (1|island),
                        data = penguins)

summary(mod3p)

```

**Functions with same name from two packages:** sometimes multiple packages use
the same names for functions that may or may not do the exactly same thing. When
loading packages you may get warnings about functions being 'masked'. This could
potentially cause quite a lot of issues. It is good practice to use
`packageName::function`, when you think that there may be multiple functions of
the same name, to make sure that your code is doing what you think it is doing.

# Non-linear relationships: polynomials and GAMs

We will look at the number of death from larynx cancer across regions in Germany
between 1986 and 1990 and levels of smoking.

```{r poly}

data(larynx)
 
mod4 <- glm(Y ~ x, family = "poisson", data = Larynx)

summary(mod4)


mod5 <- glm(Y ~ poly(x, 2), data = Larynx)

summary(mod5)

```

Polynomials are global functions, so they do not offer much flexibility.
Sometimes this may be a good thing - too much flexibility increases the risk of
overfitting.

The formula syntax for GAMs is the same as for LMs and GLMs. To specify a smooth
term, we use the `s()` function within the formula.

You can fit different types of basis (smooth). This is specified using the `bs`
argument within `s()`. By default, thin plate regression splines are used. What
to use depends on the data: e.g. cyclic cubic splies (`bs = "cc"`) for temporal
data, or even smooth random effects (`bs="re"`). See `?smooth.terms` for more
information.

You can control the "wiggliness" by specifying the number of dimensions of the
basis to be used. This is done using the `k` argument within `s()`. The optimal
smoothing is estimated, and the method to do so can be specified using the
`method` argument in `gam()`.

Similarly to GLMs, we can specify a `family` argument

```{r GAM}

mod6 <- mgcv::gam(Y ~ s(x, bs = "tp"), family = "poisson",
                  data = Larynx)

summary(mod6)


```

EDF (effective degrees of freedom) gives us (kind of) the number of basis used
to produce the smooth. An EDF = 1 indicates that the relationships is linear.

```{r GAM plotting, fig.width = 6, fig.height = 6}

plot(mod6) # default plot.gam() 

pred <- ggpredict(mod6, terms = "x")

ggplot(data = pred, aes(x = x, y = predicted)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), 
              fill = "grey50", alpha = 0.2) +
  labs(x = "Smoking level", y = "Predicted camcer deaths") +
  theme_classic()


```

GAMs are also very useful when working in two-dimensions. For example, if we
have survey data and we want to predict species occurrences spatially.

Below is an oversimplified example using sample data from the Portuguese Atlas
of Breeding Birds. This is a systematic survey of 2x2 km squares. The dataset
contains occurrences of crested larks or linnets across all surveyed tetrads. We
will focus on crested lark:

```{r GAM spatial, fig.width = 12, fig.height = 8}

# install.package("gamair")

data("bird")

brd <- bird[!is.na(bird$crestlark), ]

brd$crestlark <- as.factor(brd$crestlark)

g1 <- ggplot(data = brd, aes(x = x, y = y, colour = crestlark)) +
  geom_point() +
  scale_color_manual(values = c("#DC3220", "#005AB5")) +
    theme(axis.line = element_line(color = "black"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        panel.background = element_blank(),
        plot.background = element_rect(fill = NULL,
                                       color = NA_character_),
        panel.grid = element_line(color = "grey50"),
        plot.margin = unit(c(0, 0, 0, 0), "pt"),
        legend.key.width = unit(1, "cm"),
        legend.key.height = unit(1, "cm"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "bottom")

mod7 <- gam(crestlark ~ s(x,y), family = "binomial", data = brd)


brd$pred <- predict(mod7, newdata = brd, type = "response")

g2 <- ggplot(data = brd, aes(x = x, y = y, colour = pred)) +
  geom_point() +
  scale_color_viridis_c(option = "D", direction = 1) +
  theme(axis.line = element_line(color = "black"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        panel.background = element_blank(),
        plot.background = element_rect(fill = NULL,
                                       color = NA_character_),
        panel.grid = element_line(color = "grey50"),
        plot.margin = unit(c(0, 0, 0, 0), "pt"),
        legend.key.width = unit(1, "cm"),
        legend.key.height = unit(0.5, "cm"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.position = "bottom")

  
ggpubr::ggarrange(g1, g2, nrow = 1, ncol = 2)


```
