---
title: "Advanced statistics in R"
author: "Teresa Vaz Martins and Karolina Zalewska"
output: 
  html_document:
    theme: flatly 
    code_folding: show
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 80
    canonical: true
date: ""
---

```{css, echo=FALSE}

.lightgrey {
background-color:#F6F6F6;
color: #000; 
font-size:16px; 
text-align:left;
vertical-align: middle;
border-radius: 5px;
padding: 20px; 
80px

}

body{
font-family: Helvetica;
font-size: 14pt;
}


```

# Dimensionality reduction

## Principal component analysis (PCA)

To visualize high-dimensional data, we use PCA to map data to lower dimensions.
The method is particularly useful when the variables within the data set are
highly correlated. Due to this redundancy, PCA can be used to reduce the
original variables into a smaller number of new variables ( = *principal
components*). These will not be a reduced portion of the original variables, it
will be different variables!

We can do PCA in R in different ways, for instance

\- `prcomp`() [built-in R `stats` package],

\- `PCA`() [`FactoMineR` package, useful for visualisation],

**The basic syntax is:**

::: {style="callout:default"}
prcomp(x, ...)
:::

Arguments for `prcomp()`:

-   x: a numeric matrix or data frame

::: {style="callout:warning"}
**Note:** run `?prcomp` on the console to check for other possible arguments,
besides `x`

Of the possible arguments, note that `scale` is set as FALSE. This is a logical
value indicating whether the variables should be scaled to have unit variance
before the analysis takes place.
:::

However, it is recommended to set the argument `scale=TRUE` to standardise the
input data so that it has zero mean and variance one before doing PCA. This is
particularly recommended when variables are measured in different scales (e.g:
kilograms, grams).

Therefore, we will set the `scale` parameter to TRUE.

::: {style="callout:default"}
prcomp(x, scale = TRUE)
:::

::: {style="callout:primary"}
**Practice:** run PCA on 4 columns of the iris data.

\- which four columns? (Hint: run `?prcomp)` **x:**a numeric or complex matrix
(or data frame) which provides the data for the principal components analysis.)

\- replace \`---\` below by the appropriate terms.
:::

```{r, error=TRUE}


#' focus on the four columns
iris_4 <- iris_orig %>%
  select(- ___) #' which column to deselect?

#' do PCA
iris_pca <- prcomp( --- , scale = ---)

```

::: {style="callout:default"}
**Solution:** click Show button
:::

```{r, class.source = 'fold-hide'}

#' (do ?prcomp in the console), data x (eg iris_orig) must be numeric. Use `str` to see the type of each column: Species is a factor 

iris_orig %>% str()

#' focus on the first four columns, which are numeric
iris_4 <- iris_orig %>%
  select(-Species) #' deselect (using minus) non-numeric Species

#' do PCA on the numeric columns. Don't forget scale=TRUE
iris_pca <- prcomp(iris_4, scale = TRUE)

# Look at the results.
iris_pca 

```

### Visualise results

Although the default `stats` package includes options to produce many PCA
visualisations, I find it easier to customise `factoextra`. We will run just a
few examples, but you can find other options online, for instance here:

[PCA - Principal Component Analysis Essentials - Articles
-STHDA](%5Bhttp://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/)](<http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/>))

#### Scree plot

Visualizing the variance explained by each component helps us to visually
identify how many principal components are needed.

The percentage of variances captured by each of the new coordinates,

```{r}

var_explained = iris_pca$sdev^2 / sum(iris_pca$sdev^2)

#create scree plot
qplot(c(1:4), var_explained) + 
  geom_line() 

#calculate total variance explained by each principal component

factoextra::fviz_eig(iris_pca, addlabels = TRUE, ylim = c(0, 50))

#'?screeplot

```

```{r}

fviz_pca_ind(iris_pca ,
             geom.ind = "point", # show points
             col.ind = iris$Species, # color by species, iris$Species
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Groups")
```

--------------------------------------------------------------------------------

#### Variable correlation plot

In this plot:

\- Positively correlated variables are grouped together: Petal length and Petal
width

\- Negatively correlated variables are positioned on opposite sides of the plot
origin (opposed quadrants): sepal width and petal length?

```{r}

fviz_pca_var(iris_pca, col.var = "black")
```

We can do a correlation plot to check if the interpretation of the plot above
sounds right: petal length is very positively correlated with petal width and
anticorrelated with sepal width.

```{r}
#' correlation
M <- cor(iris[,-5])

corrplot::corrplot(M, method = 'ellipse', type = 'lower', diag = FALSE)
```

We are not investigating correlations between features in the same flower. In
this case, the results reflect the fact that Iris Virginica often have longer
petal length and width than Iris setosa, but Iris setosa often have the greatest
sepal width. The same result is confirmed looking at the column Species
(boxplots) of the first plot in the ANOVA section.

The function \`fviz_contrib()\` can be used to draw a bar plot of variable
contributions to the principal components.

```{r}
# Contributions of variables to PC1

fviz_contrib(iris_pca, choice = "var", axes = 1)

# Contributions of variables to PC2
fviz_contrib(iris_pca, choice = "var", axes = 2)
```

::: {style="callout:warning"}
**Note:** we have been using the default method for `prcomp`. But you may see a
formula being used instead of data. See documentation `?prcomp`

For instance, PCA on penguins dataset may look like:
:::

```{r}

penguins_visl <- penguins_orig %>%
  mutate(species = as.factor(species),
         island = as.factor(island),
         sex = as.factor(substr(sex,1,1))) %>%
  na.omit()


#' Notice the formula: ~bill_length_mm + bill_depth_mm + ...
penguins_pca <- prcomp(~ bill_length_mm + bill_depth_mm + 
                               flipper_length_mm + body_mass_g,
                             data=penguins_visl,
                             na.action=na.omit,  
                             scale. = TRUE)



penguins_pca
```

--------------------------------------------------------------------------------

**Further reading:**

PCA using the
[recipes](%5Bhttps://recipes.tidymodels.org/)](<https://recipes.tidymodels.org/>))
package from

[tidymodels](%5Bhttps://www.tidymodels.org/)](<https://www.tidymodels.org/>))

[PCA with penguins and recipes •
palmerpenguins(allisonhorst.github.io)](%5Bhttps://allisonhorst.github.io/palmerpenguins/articles/pca.html)](<https://allisonhorst.github.io/palmerpenguins/articles/pca.html>))

Using the factoextra package

[PCA - Principal Component Analysis Essentials - Articles
-STHDA](%5Bhttp://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/)](<http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/>))

[PCA with tidyverse - Benjamin
Nowak(bjnnowak.netlify.app)](%5Bhttps://bjnnowak.netlify.app/2021/09/15/r-pca-with-tidyverse/)](<https://bjnnowak.netlify.app/2021/09/15/r-pca-with-tidyverse/>))

# Going beyond the Gaussian error distribution: Generalized linear models (GLM)

These can be used to model data that is not normally distributed. The different
types are based on the probability distribution of the response variable: eg
Poisson if the response variable represents counts or binomial for a binary
outcome.

::: {style="callout:default"}
\`glm\` Used to fit generalized linear models with the following syntax:

glm(formula, family , data, ...)
:::

formula and data: same as in the linear model \`lm\`

Additional parameter

**family:** The statistical family to use to fit the model. Default is gaussian,
as in a linear model, but other options include binomial, Gamma, and poisson
among others. Examples:

::: {style="callout:default"}
glm(y \~ model, data = data, \*\*\`family = "binomial"\`\*\*)

glm(y \~ model, data = data, \*\*\`family = "poisson"\`\*\*)
:::

## I have count data / I'm interested in the rate : Poisson distribution

## Logistic regression

\> **Question:** which variables significantly contribute to determining the
gender of penguins.

Before beginning to build the model, double-check that all qualitative variables
have been saved as factors.

::: {style="callout:primary"}
**Practice:** Fill the \`---\` below: what is the outcome variable?
:::

```{r fitglm, error=TRUE, message=FALSE}

fit <- glm(--- ~ species + island + bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g, data = penguins_omit, family="binomial")

#' view summary of the model

#summary(fit)

```

::: {style="callout:default"}
**Solution:** Click Show button
:::

```{r, class.source = 'fold-hide'}

fit <- glm(sex ~ species + island + bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g, data = penguins_omit, family="binomial")

#' view summary of the model

summary(fit)

```

\`The "Estimate" column in the output above corresponds to the log-odds.

The odd ratios can be obtained as follows:

```{r, error=TRUE, message=FALSE}

#calculate odds ratio and 95% confidence interval for each predictor variable 

exp(cbind(Odds_Ratio = coef(fit), confint(fit)))
```

The odds ratio for each coefficient represents the average increase in the odds
of an individual defaulting, assuming all other predictor variables are held
constant.

# Mixed Effects Models in R with lme4

Mixed effects models, also known as multilevel models or hierarchical models,
are statistical models that are used to analyze data with nested or hierarchical
structures. These models are particularly useful when dealing with data that has
multiple levels of variation or when there is a need to account for the
correlation between observations within the same group.

In R, the \`lme4\` package provides functions for fitting mixed effects models.
The \`lmer()\` function is commonly used to fit linear mixed effects models

::: lightgrey
model \<- lmer(y \~ a + b + \*\*(1\|group)\*\*, data = data)
:::

Run \`?lmer\` to understand the syntax:

-   `y` is the outcome variable,

-   `a` and `b` are the fixed effects variables,

-   `group` is the random effects variable.

-   The `(1 | group)` term specifies that the random effects are intercepts that
    vary across different levels of the \`group\` variable.

::: {style="callout:primary"}
**Practice:** Analyze the effect of different diets on chick growth, using the
in-built dataset `chick_orig <- datasets::ChickWeight`.

Given the following model, could you comment the code, filling \`---\` with the
right term or terms:
:::

```{r, error=TRUE, warning=FALSE, message=FALSE}

#library(lme4) # already loaded at start

m <- lmer(weight ~ Time + Diet + Time:Diet + (1 + Time | Chick), data = chick_orig)

# formula: weight ~ Time + Diet + Time:Diet + (1 + Time | Chick)

# fixed effects: --- 

# random effects: ---

#summary(m)

```

The random effects terms `(1 + Time | Chick)` allows individual chicks to vary
randomly in terms of their intercept (starting weight) and their effect of Time
(weight change over time.

# Non-linear relationships: GAM
