---
title: "Best Coding Practices"
Author: "Ellie Harris"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: 72
---

# What are best coding practices?

We use best coding practices to help make our R code easier to **read**,
**write**, **share**, and **verify**. The [Google style
guide](https://google.github.io/styleguide/Rguide.html), and [Tidyverse
Style guide](https://style.tidyverse.org/) are a good place to start for
general rules of thumb and common preferences. The tidyverse guide was
created from Google's initial style guide, whilst Google's new style
guide is based on the tidyverse style guide, so they are very similar!

# Directory set up and R projects

<https://kdestasio.github.io/post/r_best_practices/#fn1>

The following code and examples are from Hadley Wickham's tidyverse
style guide.

# Files

## Names

File names should be meaningful and end in `.R`. Avoid using special
characters in file names - stick with numbers, letters, `-`, and `_`.

    # Good
    fit_models.R
    utility_functions.R

    # Bad
    fit models.R
    foo.r
    stuff.r

If files should be run in a particular order, prefix them with numbers.
If it seems likely you'll have more than 10 files, left pad with zero:

    00_download.R
    01_explore.R
    ...
    09_model.R
    10_visualize.R

Pay attention to capitalization, since you, or some of your
collaborators, might be using an operating system with a
case-insensitive file system (e.g., Microsoft Windows or OS X) which can
lead to problems with (case-sensitive) revision control systems. Prefer
file names that are all lower case, and never have names that differ
only in their capitalization.

## Organisation

It's hard to describe exactly how you should organise your code across
multiple files. I think the best rule of thumb is that if you can give a
file a concise name that still evokes its contents, you've arrived at a
good organisation. But getting to that point is hard.

## Internal structure

Use commented lines of `-` and `=` to break up your file into easily
readable chunks. This will also create subheadings that help to navigate
to precise points in your code.

```{r Headings, eval = FALSE}
# Load data ---------------------------

# Plot data ---------------------------
```

If your script uses add-on packages, load them all at once at the very
beginning of the file. This is more transparent than sprinkling
`library()` calls throughout your code or having hidden dependencies
that are loaded in a start-up file, such as `.Rprofile`.

```{r Loading packages, error=TRUE, echo=TRUE}
# Packages ----------------------
#install.packages("terra")
library(terra)

# see package conflicts
conflicts()  

# set package preference for conflicted functions
conflicted::conflict_prefer("filter", "dplyr")

```

# Syntax

## Object names

Variable and function names should use only lowercase letters, numbers,
and `_`. Use underscores (`_`) (so called snake case) to separate words
within a name.

```{r, eval = FALSE}
# Good
day_one
day_1

# Bad
DayOne
dayone
```

Base R uses dots in function names (`contrib.url()`) and class names
(`data.frame`), but it's better to reserve dots exclusively for the S3
object system. In S3, methods are given the name `function.class`; if
you also use `.` in function and class names, you end up with confusing
methods like `as.data.frame.data.frame()`.

If you find yourself attempting to cram data into variable names (e.g.
`model_2018`, `model_2019`, `model_2020`), consider using a list or data
frame instead.

Generally, variable names should be nouns and function names should be
verbs. Strive for names that are concise and meaningful (this is not
easy!).

```{r, eval = FALSE}
# Good
day_one

# Bad
first_day_of_the_month
djm1
```

Where possible, avoid re-using names of common functions and variables.
This will cause confusion for the readers of your code.

```{r, eval = FALSE}
# Bad
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```

## Spacing

### Commas

Always put a space after a comma, never before, just like in regular
English.

```{r, eval = FALSE}
# Good
x[, 1]

# Bad
x[,1]
x[ ,1]
x[ , 1]
```

### Parentheses

Do not put spaces inside or outside parentheses for regular function
calls.

```{r, eval = FALSE}
# Good
mean(x, na.rm = TRUE)

# Bad
mean (x, na.rm = TRUE)
mean( x, na.rm = TRUE )
```

Place a space before and after `()` when used with `if`, `for`, or
`while`.

```{r, eval = FALSE}
# Good
if (debug) {
  show(x)
}

# Bad
if(debug){
  show(x)
}
```

Place a space after `()` used for function arguments:

```{r, eval = FALSE}
# Good
function(x) {}

# Bad
function (x) {}
function(x){}
```

### Infix operators

Most infix operators (`==`, `+`, `-`, `<-`, etc.) should always be
surrounded by spaces:

```{r, eval = FALSE}
# Good
height <- (feet * 12) + inches
mean(x, na.rm = TRUE)

# Bad
height<-feet*12+inches
mean(x, na.rm=TRUE)
```

There are a few exceptions, which should never be surrounded by spaces:

-   The operators with [high
    precedence](https://rdrr.io/r/base/Syntax.html): `::`, `:::`, `$`,
    `@`, `[`, `[[`, `^`, unary `-`, unary `+`, and `:`.

    ```{r, eval = FALSE}
    # Good
    sqrt(x^2 + y^2)
    df$z
    x <- 1:10

    # Bad
    sqrt(x ^ 2 + y ^ 2)
    df $ z
    x <- 1 : 10
    ```

-   The help operator

    ```{r, eval = FALSE}
    # Good
    package?stats
    ?mean

    # Bad
    package ? stats
    ? mean
    ```

### Extra spaces

Adding extra spaces is ok if it improves alignment of `=` or `<-`.

```{r, eval = FALSE}
# Good
list(
  total = a + b + c,
  mean  = (a + b + c) / n
)

# Also fine
list(
  total = a + b + c,
  mean = (a + b + c) / n
)
```

Do not add extra spaces to places where space is not usually allowed.

## Function calls

### Named arguments {#argument-names}

A function's arguments typically fall into two broad categories: one
supplies the **data** to compute on; the other controls the **details**
of computation. When you call a function, you typically omit the names
of data arguments, because they are used so commonly. If you override
the default value of an argument, use the full name:

```{r, eval = FALSE}
# Good
mean(1:10, na.rm = TRUE)

# Bad
mean(x = 1:10, , FALSE)
mean(, TRUE, x = c(1:10, NA))
```

Avoid partial matching.

### Assignment

Avoid assignment in function calls:

```{r, eval = FALSE}
# Good
x <- complicated_function()
if (nzchar(x) < 1) {
  # do something
}

# Bad
if (nzchar(x <- complicated_function()) < 1) {
  # do something
}
```

## Control flow

### Code blocks {#indenting}

Curly braces, `{}`, define the most important hierarchy of R code. To
make this hierarchy easy to see:

-   `{` should be the last character on the line. Related code (e.g., an
    `if` clause, a function declaration, a trailing comma, ...) must be
    on the same line as the opening brace.

-   The contents should be indented by two spaces.

-   `}` should be the first character on the line.

```{r, eval = FALSE}
# Good
if (y < 0 && debug) {
  message("y is negative")
}

# Bad
if (y < 0 && debug) {
message("Y is negative")
}


```

## Long lines

Strive to limit your code to 80 characters per line. This fits
comfortably on a printed page with a reasonably sized font. If you find
yourself running out of room, this is a good indication that you should
encapsulate some of the work in a separate function.

If a function call is too long to fit on a single line, use one line
each for the function name, each argument, and the closing `)`. This
makes the code easier to read and to change later.

```{r, eval = FALSE}
# Good
do_something_very_complicated(
  something = "that",
  requires = many,
  arguments = "some of which may be long"
)

# Bad
do_something_very_complicated("that", requires, many, arguments,
                              "some of which may be long"
                              )
```

As described under [Named arguments](#argument-names), you can omit the
argument names for very common arguments (i.e. for arguments that are
used in almost every invocation of the function). Short unnamed
arguments can also go on the same line as the function name, even if the
whole function call spans multiple lines.

```{r, eval = FALSE}
map(x, f,
  extra_argument_a = 10,
  extra_argument_b = c(1, 43, 390, 210209)
)
```

You may also place several arguments on the same line if they are
closely related to each other, e.g., strings in calls to `paste()` or
`stop()`. When building strings, where possible match one line of code
to one line of output.

```{r, eval = FALSE}
# Good
paste0(
  "Requirement: ", requires, "\n",
  "Result: ", result, "\n"
)

# Bad
paste0(
  "Requirement: ", requires,
  "\n", "Result: ",
  result, "\n")
```

## Semicolons

Don't put `;` at the end of a line, and don't use `;` to put multiple
commands on one line.

## Assignment

Use `<-`, not `=`, for assignment.

```{r}
# Good
x <- 5

# Bad
x = 5
```

## Data

### Character vectors

Use `"`, not `'`, for quoting text. The only exception is when the text
already contains double quotes and no single quotes.

```{r, eval=FALSE}
# Good
"Text"
'Text with "quotes"'
'<a href="http://style.tidyverse.org">A link</a>'

# Bad
'Text'
'Text with "double" and \'single\' quotes'
```

### Logical vectors

Prefer `TRUE` and `FALSE` over `T` and `F`.

## Comments

Each line of a comment should begin with the comment symbol and a single
space: `#`

In data analysis code, use comments to record important findings and
analysis decisions. If you need comments to explain what your code is
doing, consider rewriting your code to be clearer. If you discover that
you have more comments than code, consider switching to [R
Markdown](https://rmarkdown.rstudio.com/).

In code, use comments to explain the "why" not the "what" or "how". Each
line of a comment should begin with the comment symbol and a single
space: `#`.

```{r, eval = FALSE}
# Good

# Objects like data frames are treated as leaves
x <- map_if(x, is_bare_list, recurse)


# Bad

# Recurse only with bare lists
x <- map_if(x, is_bare_list, recurse)
```

Comments should be in sentence case, and only end with a full stop if
they contain at least two sentences:

```{r, eval = FALSE}
# Good

# Objects like data frames are treated as leaves
x <- map_if(x, is_bare_list, recurse)

# Do not use `is.list()`. Objects like data frames must be treated
# as leaves.
x <- map_if(x, is_bare_list, recurse)


# Bad

# objects like data frames are treated as leaves
x <- map_if(x, is_bare_list, recurse)

# Objects like data frames are treated as leaves.
x <- map_if(x, is_bare_list, recurse)
```

# Pipes

## Introduction

Use `%>%` (tidyverse, covered later) or \|\> (base R) to emphasise a
sequence of actions, rather than the object that the actions are being
performed on.

Avoid using the pipe when:

-   You need to manipulate more than one object at a time. Reserve pipes
    for a sequence of steps applied to one primary object.

-   There are meaningful intermediate objects that could be given
    informative names.

This will be covered in more detail in the tidyverse section of the
course, but here are a few small notes on the topic:

## Whitespace

\`\` should always have a space before it, and should usually be
followed by a new line. After the first step, each line should be
indented by two spaces. This structure makes it easier to add new steps
(or rearrange existing steps) and harder to overlook a step.

```{r, eval = FALSE}
# Good
iris %>%
  group_by(Species) %>%
  summarize_if(is.numeric, mean) %>%
  ungroup() %>%
  gather(measure, value, -Species) %>%
  arrange(value)

# Bad
iris %>% group_by(Species) %>% summarize_all(mean) %>%
ungroup %>% gather(measure, value, -Species) %>%
arrange(value)
```

## Long lines

If the arguments to a function don't all fit on one line, put each
argument on its own line and indent:

```{r, eval = FALSE}
iris %>%
  group_by(Species) %>%
  summarise(
    Sepal.Length = mean(Sepal.Length),
    Sepal.Width = mean(Sepal.Width),
    Species = n_distinct(Species)
  )
```

## Assignment

There are several acceptable forms of assignment:

-   Variable name and assignment on separate lines:

    ```{r, eval = FALSE}
    iris_long <-
      iris %>%
      gather(measure, value, -Species) %>%
      arrange(-value)
    ```

-   Variable name and assignment on the same line:

    ```{r, eval = FALSE}
    iris_long <- iris %>%
      gather(measure, value, -Species) %>%
      arrange(-value)
    ```

# Useful packages

Here are some useful packages for best coding practices:

-   [styler](http://styler.r-lib.org) allows you to interactively
    restyle selected text, files, or entire projects. It includes an
    RStudio add-in, the easiest way to re-style existing code.

-   [lintr](https://github.com/jimhester/lintr) performs automated
    checks to confirm that you conform to the style guide.
