---
title: "Week 1, Day 1: Basic Data Wrangling"
author: "Kate Collingridge and Kathrin Vossen"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE) ## have set this to FALSE for now, might have to change that if we want to show outputs
Sys.setenv(tz="UTC") #set environment time zone to utc

```

## Basic data wrangling

In this session, we will take a look at functions and packages that will help you do some basic data exploration. We will investigate some example data, take a look at pivoting data, learn how to merge two datasets and create subsets of data. All the data we will be using are example datasets that are built into R, so there is no need to import external datasets.

We will need to make sure we have installed and loaded the following packages:

```{r packages}
library(tidyverse)
library(reshape2)
# install.packages("ViewPipeSteps")
library(ViewPipeSteps)
#we might need to explain here if teresa hasn't already what all the warnings you get when you load tidyverse are
```


### Investigating your data
Let's start by taking a look at the `iris` dataset:

```{r investigate1}
data("iris")

# It can be useful to make a copy of your original dataset. That way, if something goes wrong, you can go back to a previous version of your data.

iris_ori <- iris

# Let's get a first overview

str(iris)

# We can see that we have 4 columns containing numerical values and one column containing a factor with 3 levels

summary(iris)
head(iris)
tail(iris)

## Let's take a closer look at the rows and columns in our dataset:

ncol(iris) # we have 5 columns
nrow(iris) # we have 150 rows

# What are minimum and maximum values for our numerical columns?
min(iris$Sepal.Length)
max(iris$Sepal.Width)
mean(iris$Petal.Length)

# If we wanted we could calculate the sum of a column:
sum(iris$Sepal.Length)

## Now we want to take a look at our factors - what are the three levels?
unique(iris$Species)

# We can use this function on any column, but with numerical values we will just get more results:
unique(iris$Sepal.Length)

# Do we have any duplicates in our dataset? Do we have any NAs?
any(is.na(iris$Sepal.Length))
table(is.na(iris$Sepal.Length))

any(duplicated(iris$Sepal.Length))
duplicated(iris$Sepal.Length)
table(duplicated(iris$Sepal.Length))

# In this case this is not very informative -- we expect to see 'duplicate' entries
```

### Making basic changes to your data
We can change things such as column names, or set them to all lower or upper case. This can be useful when we want to merge different datasets together.

```{r change}
# let's change a column name
names(iris)[names(iris)=="Species"] <- "Type" 
str(iris)

# let's replace dots in column names with underscores

names(iris) <- gsub("\\.", "_", names(iris))
str(iris)

# what if we want to change column names or rows to all upper or all lower case

names(iris) <- toupper(names(iris))
str(iris)

names(iris)[names(iris)=="TYPE"] <- tolower(names(iris)[names(iris)=="TYPE"])
str(iris)

#we can also change all values in a column
iris$type <- toupper(iris$type)
str(iris)

# let's reverse these changes for now
iris <- iris_ori

```



- structure of your data -- Sam covers summary() -- done
- number of rows, columns; getting the sum of a column -- done
- unique values -- done
- how to change column names -- done
- how to set column names and values to all lower or all upper case -- done
- how to sort vectors (using base R sort function) -- **needs to be added**

### Expanding vs reducing your datasets
Now that we have explored the `iris` data a bit, let's try subsetting it based on different conditions. First, let's try making a new dataframe where we only select data for the species "setosa". There are two ways you can do this - using R base functions or using a Tidyverse pipeline.

Let's try a conditional statement using square brackets first:
```{r subset1}
setosa <- iris[iris$Species == "setosa",]
str(setosa)
unique(setosa$Species) # Note R still knows there are 3 levels for this factor, but the dataset only includes 1
```

You can read this statement as: "assign an object called 'setosa' using the 'iris' dataset, selecting all rows where the factor Species has the value 'setosa', and selecting all columns". We are selecting **rows** because our conditional statement is *before* the comma.

We can select **columns** if the conditional statement is *after* the comma. For example, we can only select two of the columns:

```{r subset2}
# to select the first three columns:
subset1 <- iris[,1:3]
str(subset1)
# or to select by column name:
subset2 <- iris[,c("Sepal.Length", "Sepal.Width")]
str(subset2)
```

Now let's try the same using a Tidyverse function. First, this is how to select specific **rows** based on a factor:

```{r subset3}
setosa <- iris %>%
  filter(Species == "setosa")
str(setosa)
```

And this is how to select specific **columns**:

```{r subset4}
# select first three columns:
subset3 <- iris %>%
  select(1:3)

# select by column name:
subset4 <- iris %>%
  select(Sepal.Length, Sepal.Width)
```


The next thing we want to try is expanding our dataset by adding a new column, Let's say we want to calculate the petal length to width ratio. We can do this calculation and automatically save this in a new column that we will call "LWRatio":

```{r column1}
iris$LWRatio <- iris$Petal.Length / iris$Petal.Width
```

We saw before that we can check data for duplicate entries. In the case of the `iris` dataset, we expect to see duplicate entries when we look at individual columns -- we would assume that the same petal length or width will show up multiple times. However, sometimes we need to check if we have true duplicates in our data. We can do this by creating a 'unique identifier' and then using this column to check for duplicates.

```{r duplicates}
# let's create a unique identifier by pasting together information from the different rows: (KC think we could have a better example here)

iris$UniqueID <- apply(iris[,1:6], 1, paste, collapse = "-")
str(iris)
# now check
table(duplicated(iris$UniqueID))


# we have one duplicate entry
# let's check that further

ID <- iris$UniqueID[duplicated(iris$UniqueID) == TRUE]
test <- iris[iris$UniqueID == ID,]

# we can remove one of the two entries using the row number:
iris <- iris[-143,]

#An alternative way to do this is by using dplyr package:

dup <- iris %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()

dup

#Keep only the distinct rows (based on all columns)
iris <- iris %>% distinct()
nrow(iris) #we now have only 149 rows, we have lost the duplicate.


#What about if you have a more complex dataset, where you might have duplicate samples and want to keep only one of them? For example, say you have combined two datasets, but they contain some of the same data and you only want one version of each sample.

```


- adding new columns -- done
- checking for duplicates -- done, want to add more complex example though.
- summing across rows and columns -- **do we want to include this?**
- creating subsets: filtering and subsetting using conditional statements (aka base R functions) -- done

### Wide vs long format

There are two main ways to store datasets - wide and long. Datasets are usually easier to work with in R in long format, but you may need to switch between them

Long format is where you have data in an identifying column (or columns) which repeats, and then your response data all in one column. For example, let's look at the chick weight dataset.

```{r long}
#oad the chick weight dataset
data("ChickWeight")
ChickWeight <- as.data.frame(ChickWeight)
str(ChickWeight)
#reorder the columns into a more sensible order

ChickWeight <- ChickWeight[, c("Chick", "Diet", "Time", "weight")] # leave the row index blank to keep all rows

View(ChickWeight)
nrow(ChickWeight)
unique(ChickWeight$Chick)
```

You can see that this dataset has four columns. The first 3 contain identifying info (which chick is it and which diet is it on) and time of the measurement. The final column has the chick weight. Thus there are only 50 different chicks on 4 diets, but as they were weighed at different times and the data is in long format there are 578 rows in total. This is a good format for many r functions, but it does repeat information and is not so easy to read.

The other way to store this data is in wide format, where each chick just has one row, and each time it is weighed goes in a separate column

```{r wide}
#In tidyverse
ChickWeight %>% tidyr::spread(Time, weight)
#In reshape2
reshape2::dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight")
```

There are various terms for converting between these formats. The most common are melt or gather for wide>long, and cast or spread for long>wide.

Casting a dataset is simple where each unique entity only has one of each measurement, but sometimes that isn't the case. Maybe it was a bit difficult to get the chicks to stay still for weighing and you actually had more than one measurement for each chick in order to take an average. If you tried to cast such a dataset then you would get an error, as it can't put more than one weight in each cell in the wide format table. In this case you might need to specify a function, usually average. 

```{r wide2, error = TRUE}
ChickWeight

#let's say chick 1 was being particularly difficult to weigh on day 6, so we have multiple measurements for that day
ChickWeight <- ChickWeight %>% add_row(Chick = "1", Diet = "1", Time = 6, weight = 66)

ChickWeight %>% tidyr::spread(Time, weight)
#See how we get an error this time, because row 4 and 579 both contain weights for chick 1, diet 1, day 6.

reshape2::dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight")
#In dcast it does run but says "Aggregation function missing: defaulting to length", and you will see all but one cells have a length of 1.
#If we add fun.aggregate = mean the error goes away and where there are duplicates it calculates an average

ChickWeight_wide <- dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight", fun.aggregate = mean)

```
This method can also be used to do a group by query which we will look at in more detail later.

There may also be situations where you want to go from wide to long format

```{r wide to long}

ChickWeight_long <- reshape2::melt(ChickWeight_wide, id = c("Chick", "Diet"), value.name = "weight", variable.name = "Time")
```
Now we have the data back in long form. However you may notice that there are now 600 observations, when there were originally 578. This is because each cell has been given a row regardless of whether there was a measurement in it, so there are 50 chicks * 12 times = 600 rows. As you may have noticed in the wide table, several of the chicks weren't measured every time step (let's imagine they escaped part way through the experiment), for example chick 16 was only weighed up to day 12. So by pivoting from wide to long format we have introduced some extra rows.

```{r remove NAs}
ChickWeight_long <- ChickWeight_long[!is.na(ChickWeight_long$weight),]
```
Now we are back to the orginal 578 rows.


### Merging datasets
Before we can merge or join two datasets together, we need to check that they are compatible. The `janitor` package allows us to compare columns between two datasets. Let's create an example by setting up two subsets of the `iris` dataset and renaming one column in one of the subsets to create a difference:

```{r janitor, eval = TRUE, warning=FALSE, message=FALSE}

# iris <- iris1
subset1 <- iris[iris$Species == "setosa",]
subset2 <- iris[iris$Species == "virginica",]

names(subset2)[names(subset2) == "Species"] <- "Type"

library(janitor)

janitor::compare_df_cols(subset1, subset2)
```
To join two datasets together, we have several options:

- **cbind**: this function merges two dataframes by adding *columns* -- so the dataframe is expanded horizontally
- **rbind**: this function merges two dataframes by adding *rows* -- so the dataframe is expanded vertically
- **merge**: base R function within which you can specify different ways to merge two dataframes 
- **Tidyverse options**:
  - inner_join() corresponds to merge(X, Y) – only retain rows that match between X and Y.
  - left_join() corresponds to merge(X, Y, all.x = T) – retain all rows from X, whether or not they match in Y.
  - right_join() corresponds to merge(X, Y, all.y = T) – retain all rows from Y, whether or not they match in X.
  - full_join() corresponds to merge(X, Y, all = T) – retain all rows from both X and Y, whether or not they match.

Source and further info: https://bookdown.org/rwnahhas/IntroToR/merge-join.html

An example:

```{r merge}
#set the name back to species for subset 2
names(subset2)[names(subset2) == "Type"] <- "Species"

fullDF <- rbind(subset1, subset2)

```

Now lets look at some made up data

```{r schools}
# Create some dataframes with random data.

# Random ratings per school and city.
schools <- data.frame(City=c('Lowestoft', 'Lowestoft', 'Beccles', 'Beccles', 'Norwich'),
                      School=c('Pakefield', 'Kirkley', 'John Leman', 'Free School', 'Hewitt'),
                      Rating=c('B', 'D', 'C', 'C', 'B'))

# Town sizes.
towns <- data.frame(Town=c('Lowestoft', 'Beccles', 'Norwich'),
                    Population=c(30000, 10000, 270000))

NewRating <- data.frame(NewRating=c('B', 'C', 'B', 'D', 'B'))

#Add the information about population size to the schools table using merge
schools2 <- merge(schools, towns, by.x = "City", by.y = "Town")
#Note we had to specify the column to match by as it was called city in one table and town in another. If we change the column name to be town in both then the merge is simpler as it guesses that you want to join on that column.
names(schools)[names(schools) == "City"] <- "Town"

schools2 <- merge(schools, towns)

# Suppose the schools have been rated again and we want to add the new ratings, we can use cbind


# Example of how to use cbind (column bind).
schools3 <- cbind(schools2, NewRating)

```




### Dealing with dates and times

Most data will come with a date and or time, but the formats can vary widely so it is important to understand how to deal with these in R. ["If you think dates and times are simple you haven't thought about it enough!"](https://gist.github.com/timvisee/fcda9bbdff88d45cc9061606b4b923ca)

#### Time zones

The first thing to get right is the time zone!
You may have noticed that in the first setup chunk of code we set the timezone to UTC. This is the time zone at °0 longitude and does not use daylight saving time so it is always the same. As a default it is good to set this at the beginning of your script.

```{r timezones}
Sys.setenv(tz="UTC") #set environment time zone to utc
#To view the timezone
Sys.timezone()
#You can set a differnt timezone, e.g.
Sys.setenv(tz="CET") #set environment time zone to central european time
Sys.timezone()

#Let's set it back to UTC for now
Sys.setenv(tz="UTC") 
Sys.timezone()

#We can also get R to tell us the current date and time
Sys.Date()
Sys.time()
#We can do this with the lubridate package too
lubridate::today()
lubridate::now()
#We can even get it to tell us the time in a different zone, say if I want to know if it's a good time to call my kiwi relatives.
lubridate::now(tzone = "NZ")

```
OlsonNames() will show you all the timezone names.


#### Daylight savings time

One of the complications of time is that depending on the time zone and whether the data has been recorded in local time, it may inconveniently shift by an hour for part of the year. For example if your data was collected in the UK in the summer time it may be in British Summer Time, which is UTC+1. This gets further complicated if you have data that spans the clock change. 
[timeanddate.com](timeanddate.com) gives information about when daylight savings is used in each timezone e.g. [https://www.timeanddate.com/time/zones/cet](https://www.timeanddate.com/time/zones/cet)
But I would recommend to always use UTC as your sys timezone!
If you want a bit more info about timezones read this [R blog about timezone troubles](https://www.r-bloggers.com/2018/07/a-tour-of-timezones-troubles-in-r/)

Now let's learn how to format dates and times in R

#### Formatting dates and times

Your data may store this information in many different ways. Maybe you have years, days and months in separate columns, maybe months are in numbers, maybe in words, maybe in 3 letter abbreviations. Maybe your dates are in UK format, maybe American style with the day and month switched. Maybe your datetime is all in one column, maybe date and time are separate. Perhaps you have a separate column that tells you the timezone (if so, hallelujah someone thought to include this vital info!). Times may be in 12 or 24 hour format, and may or may not include seconds
Here are some examples:

08/01/2024
2024/08/01
01/08/24
8th January 2024
08/01/2024 14:03
08/01/2024 02:03 PM
8 Jan 2024 14 03 06
08/01/2024 15:03 CET
8th January 2024
202401081T140306Z
1704722580

You can probably read and understand all of those (except the last one probably), but R needs to be told what they all are so it knows what datetime to associate with your data.

Let's take a look at datetime classes. The main ones used are Date (a day stored as the number of days since 1970-01-01), and POSIXct (date and time in seconds with the number of seconds beginning at 1 January 1970)

```{r date class}
class(Sys.Date()) # Date class
class(lubridate::now()) #POSIXct class


```
Now let's take some of the date formats above and tell R how to parse them

```{r date parsing}
date <- "2024/08/01"
class(date) # we see that R just has this stored as a string
date1 <- as.Date(date)
str(date1) # now stored as a date
#However, if we had give it a date like this
date <- "01/08/2024"
date2 <- as.Date(date)
str(date2) #you can see something has gone wrong, it has confused the year and the day. This is because the format wasn't what it expected, so we need to specify the format.
date2 <- as.Date(date, format = "%d/%m/%Y")
str(date2) # now it is correct as we told it what to expect.

#as.Date is a base R function, we can also use the very similar lubridate::as_date function

```
Exercise - can you take some of the date formats above and parse them so R understands them? Use [this guide](https://www.statology.org/r-date-format/) to understand how to specify the format.

What about when we also have times?

```{r datetime parsing}
datetime <- "08/01/2024 14:03:00"
#parse this to posixct, specifying the format
datetime1 <- as.POSIXct(datetime, format = "%d/%m/%Y %H:%M:%S", tz="UTC") #note we specify the timezone, always a good idea! But if you don't it will assume what you set as your sys.timezone
str(datetime1)


datetime <- "20240108T140306Z"
datetime2 <- as.POSIXct(datetime, format = "%Y%m%dT%H%M%SZ", tz="UTC")
str(datetime2)

#We can also see how this is stored as a number
as.numeric(datetime2)

#Which should help you with the last format 1704722586 which is actually the simplest as it is the number of seconds since 1st Jan 1970. 
datetime2 <- as.POSIXct(1704722586, origin = "1970-01-01")
str(datetime2)

#1st Jan 1970 is the standard to count from, but sometimes you get numeric date times with different origins - usually 1st Jan 1900. Just change the origin to parse these.

#Sometimes we have the information spread out, e.g.

Date = c("08/01/2024")
Time = c("14:03:00")
datetime3 <- as.POSIXct(paste(Date, Time, sep = " "), format = "%d/%m/%Y %H:%M:%S")
str(datetime3)



#If you have your information in a data frame, the parsing is pretty similar, just specify the columns


course_dates <- data.frame(Date = c("08/01/2024", "09/01/2024", "22/01/2024", "23/01/2024"),
                    StartTime = c("09:00:00", "09:00:00", "09:00:00", "09:00:00"),
                    EndTime = c("16:00:00", "16:00:00", "16:00:00", "15:00:00"),
                    TimeZone = c("UTC", "UTC", "UTC", "UTC"), 
                    Location = "Zicer", "Zicer", "", "")

course_dates$StartDateTime <- as.POSIXct(paste(course_dates$Date, course_dates$StartTime, sep = " "), format = "%d/%m/%Y %H:%M:%S", tz = "UTC")

course_dates$EndDateTime <- as.POSIXct(paste(course_dates$Date, course_dates$EndTime, sep = " "), format = "%d/%m/%Y %H:%M:%S")
course_dates$duration <- difftime( course_dates$EndDateTime, course_dates$StartDateTime)

```
Exercise - Now we have the start and end date times, can you calculate how many hours each day is? Hint use the difftime function



Once you have your dates parsed you can then extract info from them.

```{r extract date components}
#extract the year
course_dates$Year <- lubridate::year(course_dates$StartDateTime)
course_dates$Month <- lubridate::month(course_dates$StartDateTime, label = TRUE) #Label = TRUE gives it as month abbr (Jan etc), false gives it as a number

#you can also convert between formats e.g.
course_dates$Month_long <- month.name[course_dates$Month]

#You can also use the format function to extract components

course_dates$day <- format(course_dates$StartDateTime,"%d")
```



## Exercise

Read in the example dataset and create a posixct datetime field. Create a "Year" and "Month" column


### Dealing with spatial data

We now understand how to deal with temporal data, but most data we use is also spatial. 

R has become a very powerful tool to handle, process and map spatial data thanks to packages that have been developed over the years. Up until very recently, the packages most people were using to import and handle spatial data were `rgdal`, `rgeos` and `terra`. These packages are being retired and are being replaced by [`sf` - Simple Features](https://r-spatial.github.io/sf/.


```{r quakes}
library(sf)
#Let's use the quakes dataset, which contains the locations of earthquakes off Fiji
data("quakes")
str(quakes)
#You can see this data includes latitude and longitude, so it is spatial, but these are just stored as numeric, R does not know they are coordinates.
quakes_sf <- sf::st_as_sf(quakes, coords = c("long","lat")) # your x coordinate goes first, y second. Make sure to get this right or your data will be sideways!
str(quakes_sf) # we can now see it has geometry
plot(st_geometry(quakes_sf))


```

Similarly to timezones, with spatial data we have coordinate reference systems to deal with.

```{r crs}
#We can check the coordinate reference system of our spatial data like this
st_crs(quakes_sf)
#You will see that it is NA, so let's fix that
st_crs(quakes_sf) <- st_crs(4326) #4326 is the EPSG code for WGS84. This is the most commonly used lat/lon projection
st_crs(quakes_sf) 

#The other most common CRS is the British National Grid which uses eastings and northings.

# BNG example 

```

We will go into spatial data analysis in more detail in week 2.

### Reading in spatial data

You may also have data which is already in a spatial format, for example a shapefile. sf has functions to read these data types in.

(include example)

### Other -- For Kate to check

- json? netcdf etc (maybe for second week)
- assign one column based on another. qc - flagging data

### potential overlap with tidyverse session:
- duplicates/distinct
- selecting columns and rows -- easiest done as part of a pipeline
- filtering and subsetting
- pivoting to/from long and wide formats
- group by


