---
title: "Week 1, Day 1: Basic Data Wrangling"
author: "Kate Collingridge and Kathrin Vossen"
date: "`r format(Sys.time(), '%d/%m/%y')`"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = TRUE) ## have set this to FALSE for now, might have to change that if we want to show outputs
```

## Basic data wrangling
What has already been covered in previous sessions:

- how to import data
- how to load packages
- how to check your data -- format/classes etc (Sam's session)


In this session, we will take a look at functions and packages that will help you do some basic data exploration. We will investigate some example data, take a look at pivoting data, learn how to merge two datasets and create subsets of data. All the data we will be using are example datasets that are built into R, so there is no need to import external datasets.

We will need to make sure we have installed and loaded the following packages:

```{r packages}
library(tidyverse)
#we might need to explain here if teresa hasn't already what all the warnings you get when you load tidyverse are
```


### Investigating your data
Let's start by taking a look at the `iris` dataset:

```{r investigate1}
data("iris")

#It can be useful to make a copy of your original dataset. That way, if something goes wrong, you can go back to a previous version of your data.

iris_ori <- iris

# let's get a first overview -- this will have been covered to some extend by Sam

str(iris)

# we can see that we have 4 columns containing numerical values and one column containing a factor with 3 levels

summary(iris)
head(iris)

## let's take a closer look at the rows and columns in our dataset:

ncol(iris) # we have 5 columns
nrow(iris) # we have 150 rows

# what are minimum and maximum values for our numerical columns?
min(iris$Sepal.Length)
max(iris$Sepal.Width)
mean(iris$Petal.Length)

# if we wanted we could calculate the sum of a column:
sum(iris$Sepal.Length)

## now we want to take a look at our factors - what are the three levels?
unique(iris$Species)

# we can use this function on any column, but with numerical values we will just get more results:
unique(iris$Sepal.Length)

# do we have any duplicates in our dataset? Do we have any NAs?
any(is.na(iris$Sepal.Length))
table(is.na(iris$Sepal.Length))

any(duplicated(iris$Sepal.Length))
duplicated(iris$Sepal.Length)
table(duplicated(iris$Sepal.Length))

# in this case this is not very informative -- we expect to see 'duplicate' entries
```

### Making basic changes to your data
We can change things such as column names, or set them to all lower or upper case. This can be useful when we want to merge different datasets together.

```{r change}
# let's change a column name
names(iris)[names(iris)=="Species"] <- "Type" 
str(iris)

# let's replace dots in column names with underscores

names(iris) <- gsub("\\.", "_", names(iris))
str(iris)

# what if we want to change column names or rows to all upper or all lower case

names(iris) <- toupper(names(iris))
str(iris)

names(iris)[names(iris)=="TYPE"] <- tolower(names(iris)[names(iris)=="TYPE"])
str(iris)

#we can also change all values in a column
iris$type <- toupper(iris$type)
str(iris)

# let's reverse these changes for now
iris <- iris_ori

```



- structure of your data -- Sam covers summary() -- done
- number of rows, columns; getting the sum of a column -- done
- unique values -- done
- how to change column names -- done
- how to set column names and values to all lower or all upper case -- done
- how to sort vectors (using base R sort function) -- **needs to be added**

### Expanding vs reducing your datasets
Now that we have explored the `iris` data a bit, let's try subsetting it based on different conditions. First, let's try making a new dataframe where we only select data for the species "setosa". There are two ways you can do this - using R base functions or using a Tidyverse pipeline.

Let's try a conditional statement using square brackets first:
```{r subset1}
setosa <- iris[iris$Species == "setosa",]
str(setosa)
unique(setosa$Species) # Note R still knows there are 3 levels for this factor, but the dataset only includes 1
```

You can read this statement as: "assign an object called 'setosa' using the 'iris' dataset, selecting all rows where the factor Species has the value 'setosa', and selecting all columns". We are selecting **rows** because our conditional statement is *before* the comma.

We can select **columns** if the conditional statement is *after* the comma. For example, we can only select two of the columns:

```{r subset2}
# to select the first three columns:
subset1 <- iris[,1:3]
str(subset1)
# or to select by column name:
subset2 <- iris[,c("Sepal.Length", "Sepal.Width")]
str(subset2)
```

Now let's try the same using a Tidyverse function. First, this is how to select specific **rows** based on a factor:

```{r subset3}
setosa <- iris %>%
  filter(Species == "setosa")
str(setosa)
```

And this is how to select specific **columns**:

```{r subset4}
# select first three columns:
subset3 <- iris %>%
  select(1:3)

# select by column name:
subset4 <- iris %>%
  select(Sepal.Length, Sepal.Width)
```


The next thing we want to try is expanding our dataset by adding a new column, Let's say we want to calculate the petal length to width ratio. We can do this calculation and automatically save this in a new column that we will call "LWRatio":

```{r column1}
iris$LWRatio <- iris$Petal.Length / iris$Petal.Width
```

We saw before that we can check data for duplicate entries. In the case of the `iris` dataset, we expect to see duplicate entries when we look at individual columns -- we would assume that the same petal length or width will show up multiple times. However, sometimes we need to check if we have true duplicates in our data. We can do this by creating a 'unique identifier' and then using this column to check for duplicates.

```{r duplicates}
# let's create a unique identifier by pasting together information from the different rows: (KC think we could have a better example here)

iris$UniqueID <- apply(iris[,1:6], 1, paste, collapse = "-")
str(iris)
# now check
table(duplicated(iris$UniqueID))


# we have one duplicate entry
# let's check that further

ID <- iris$UniqueID[duplicated(iris$UniqueID) == TRUE]
test <- iris[iris$UniqueID == ID,]

# we can remove one of the two entries using the row number:
iris <- iris[-143,]

#An alternative way to do this is by using dplyr package:

dup <- iris %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()

dup

#Keep only the distinct rows (based on all columns)
iris <- iris %>% distinct()
nrow(iris) #we now have only 149 rows, we have lost the duplicate.


#What about if you have a more complex dataset, where you might have duplicate samples and want to keep only one of them? For example, say you have combined two datasets, but they contain some of the same data and you only want one version of each sample.

```


- adding new columns -- done
- checking for duplicates -- done, want to add more complex example though.
- summing across rows and columns -- **do we want to include this?**
- creating subsets: filtering and subsetting using conditional statements (aka base R functions) -- done

### Wide vs long format

There are two main ways to store datasets - wide and long. Datasets are usually easier to work with in R in long format, but you may need to switch between them

Long format is where you have data in an identifying column (or columns) which repeats, and then your response data all in one column. For example, let's look at the chick weight dataset.

```{r long}
#oad the chick weight dataset
data("ChickWeight")
ChickWeight <- as.data.frame(ChickWeight)
str(ChickWeight)
#reorder the columns into a more sensible order

ChickWeight <- ChickWeight[, c("Chick", "Diet", "Time", "weight")] # leave the row index blank to keep all rows

View(ChickWeight)
nrow(ChickWeight)
unique(ChickWeight$Chick)
```

You can see that this dataset has four columns. The first 3 contain identifying info (which chick is it and which diet is it on) and time of the measurement. The final column has the chick weight. Thus there are only 50 different chicks on 4 diets, but as they were weighed at different times and the data is in long format there are 578 rows in total. This is a good format for many r functions, but it does repeat information and is not so easy to read.

The other way to store this data is in wide format, where each chick just has one row, and each time it is weighed goes in a separate column

```{r wide}
#In tidyverse
ChickWeight %>% tidyr::spread(Time, weight)
#In reshape2
reshape2::dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight")
```

There are various terms for converting between these formats. The most common are melt or gather for wide>long, and cast or spread for long>wide.

Casting a dataset is simple where each unique entity only has one of each measurement, but sometimes that isn't the case. Maybe it was a bit difficult to get the chicks to stay still for weighing and you actually had more than one measurement for each chick in order to take an average. If you tried to cast such a dataset then you would get an error, as it can't put more than one weight in each cell in the wide format table. In this case you might need to specify a function, usually average. 

```{r wide2}
ChickWeight

#let's say chick 1 was being particularly difficult to weigh on day 6, so we have multiple measurements for that day
ChickWeight <- ChickWeight %>% add_row(Chick = "1", Diet = "1", Time = 6, weight = 66)

ChickWeight %>% tidyr::spread(Time, weight)
#See how we get an error this time, because row 4 and 579 both contain weights for chick 1, diet 1, day 6.

reshape2::dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight")
#In dcast it does run but says "Aggregation function missing: defaulting to length", and you will see all but one cells have a length of 1.
#If we add fun.aggregate = mean the error goes away and where there are duplicates it calculates an average

dcast(ChickWeight, Chick + Diet ~ Time, value.var = "weight", fun.aggregate = mean)

```
This method can also be used to do a group by query.

- use iris - pivot by species name
- pivoting datasets
- dealing with the introduced NAs

### Merging datasets
Before we can merge or join two datasets together, we need to check that they are compatible. The `janitor` package allows us to compare columns between two datasets. Let's create an example by setting up two subsets of the `iris` dataset and renaming one column in one of the subsets to create a difference:

```{r janitor, eval = TRUE, warning=FALSE, message=FALSE}

# iris <- iris1
subset1 <- iris[iris$Species == "setosa",]
subset2 <- iris[iris$Species == "virginica",]

names(subset2)[names(subset2) == "Species"] <- "Type"

library(janitor)

janitor::compare_df_cols(subset1, subset2)
```
To join two datasets together, we have several options:

- **cbind**: this function merges two dataframes by adding *columns* -- so the dataframe is expanded horizontally
- **rbind**: this function merges two dataframes by adding *rows* -- so the dataframe is expanded vertically
- **merge**: base R function within which you can specify different ways to merge two dataframes 
- **Tidyverse options**:
  - inner_join() corresponds to merge(X, Y) – only retain rows that match between X and Y.
  - left_join() corresponds to merge(X, Y, all.x = T) – retain all rows from X, whether or not they match in Y.
  - right_join() corresponds to merge(X, Y, all.y = T) – retain all rows from Y, whether or not they match in X.
  - full_join() corresponds to merge(X, Y, all = T) – retain all rows from both X and Y, whether or not they match.

Source and further info: https://bookdown.org/rwnahhas/IntroToR/merge-join.html

An example:

```{r merge}
names(subset2)[names(subset2) == "Type"] <- "Species"

fullDF <- rbind(subset1, subset2)

```



- janitor package -- comparing two datasets before merging -- done??
- cbind + rbind -- done
- merge - use data collected at smartbuoys as an example?


### Dealing with dates

- use airquality dataset for this? At least it includes month
- how to check what format a date column is in
- how to set a date
- how to extract information such as the year and month from a date


### Dealing with spatial data
R has become a very powerful tool to handle, process and map spatial data thanks to packages that have been developed over the years. Up until very recently, the packages most people were using to import and handle spatial data were `rgdal`, `rgeos` and `terra`. These packages are being retired and are being replaced by `sf` - Simple Features.

- use island dataset?
- converting to spatial formats? or reading in spatial data
- R built in dataset 'quakes' could be good one here - gives the location of earthquakes off Fiji

### Other -- For Kate to check

- json? netcdf etc (maybe for second week)
- assign one column based on another. qc - flagging data

### potential overlap with tidyverse session:
- duplicates/distinct
- selecting columns and rows -- easiest done as part of a pipeline
- filtering and subsetting
- pivoting to/from long and wide formats
- group by


