---
title: "week_1_content_rcourse"
author: "Kate Collingridge and Kathrin Vossen"
date: "16/11/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic data wrangling

(already covered how to import data)


changing formats/class
column names
toupper tolower
formatting dates
selecting columns and rows
number of rows
getting unique values adn detecting duplicates
sorting
long and wide format and how to convert between them. na issues
janitor package - to compare across datasets.
cbind, rbind, merge - use data collected at smartbuoys as an example?
filtering/subsetting
summary queries - group by
assign one column based on another. qc - flagging data
Summing across rows and columns
converting to spatial formats? or reading in spatial data

json? netcdf etc (maybe for second week)



```{r cars}
summary(cars)
```

## Basic data visualisation

ggplot etc, including how to export for publications, leaflet
histogram - in base plot and ggplot
basic scatter plot - in base plot and ggplot
adding labels etc
adding lines
grouping by factors
adding trendlines
saving plots
interactivity with plotly
cowplot for panels
sizing, formatting etc.

plotting spatial data - with ggplot, and with leaflet.
animations - basic gif

```{r pressure, echo=FALSE}
plot(pressure)
```

## Accessibility
https://www.jumpingrivers.com/blog/accessible-shiny-standards-wcag/
standards, why it is important
how to add alt text to plots and tables
labels
colour ramps - continuous and discrete data
https://stackoverflow.com/questions/57153428/r-plot-color-combinations-that-are-colorblind-accessible
color blindness
using symbols etc so color isn't the only way to convey info.
text size
html tags
screen readers

further reading - gov digital service, w3c
```{r pressure, echo=FALSE}
plot(pressure)
```
